---
title: "Analysis of P1 MYT1L snRNAseq"
author: "Simona Sarafinovska"
date: "2024-06-01"
---

# Set up environment
```{r dependencies, warning=FALSE, message=FALSE}
# Change this to your directory
.libPaths("/ref/jdlab/software/r-envs/simona/r-packages/4.2/")

# Load dependencies
library(dplyr)
library(Seurat)
library(scCustomize)
library(ggplot2)
library(patchwork)

set.seed(420)

source("../color_utils.R")
```

# Load in preprocessed data
```{r load}
combined_annotated <- readRDS("../../data/250308_combined_annotated_1-5_5pct_res2.rds")
combined_annotated
```
You should have a Seurat object with 138033 nuclei and 37462 features. 
# Subset for neuronal only
```{r subset neuronal}
Idents(combined_annotated) <- combined_annotated$RNA_snn_res.2

non_neuronal <- c(
  "12", # oligos
  "27", # oligos
  "41", # astros
  "48", # oligos
  "55", # oligodendrocyte precursor cells
  "56", # vasc
  "57", # astro/oligo
  "65", # oligo
  "67", # oligo
  "76", # astro
  "77", # oligo
  "78"
) # tanycytes

combined_neuronal <- subset(combined_annotated,
  idents = non_neuronal,
  invert = TRUE
)
combined_neuronal
```
You should have a Seurat object with 37462 features (genes) and 127456 samples (nuclei). 


# QC plots

QC metrics and library statistics after subsetting 

```{r genes/umis per sample}
p3 <- QC_Plots_Genes(
  seurat_object = combined_neuronal,
  group.by = "sample",
  pt.size = 0
)
p4 <- QC_Plots_UMIs(
  seurat_object = combined_neuronal,
  group.by = "sample",
  pt.size = 0
)

wrap_plots(p3, p4, ncol = 2)
```

```{r nucs per sample}
Plot_Cells_per_Sample(seurat_object = combined_neuronal, sample_col = "sample", group_by = "genotype")
```

This shows the number of nuclei per sample.
```{r table nucs per sample}
samplesTable <- table(combined_neuronal@meta.data$sample)
samplesTable
```

This shows the number of nuclei per genotype.
```{r table nucs per geno}
genotypesTable <- table(combined_neuronal@meta.data$genotype)
genotypesTable
```

## for Geno.
This creates a dataframe that summarizes the statistics across all samples and genotypes.
```{r df stats geno}
df <- data.frame(
  numCells = as.vector(samplesTable),
  genotype = c(rep(c("Het"), times = 16), rep(c("WT"), times = 20))
)

df$genotype <- factor(df$genotype, levels = c("WT", "Het"))

averages <- aggregate(numCells ~ genotype, data = df, FUN = mean)
sds <- aggregate(numCells ~ genotype, data = df, FUN = function(x) sd(x))

df2 <- merge(averages, sds, by = "genotype")
colnames(df2) <- c("genotype", "mean", "sd")

df2$genotype <- factor(df2$genotype, levels = c("WT", "Het"))
df2 <- df2[order(df2$genotype), ]

df
df2
```

We will now plot 3 plots. 
1. A barplot that shows the average number of nuclei per replicate per genotype.
2. A boxplot that shows the distribution of genes per nuclei per replicate.
3. A boxplot that shows the distribution of UMIs per nuclei per replicate.
```{r bar and boxplots geno}
p1 <- ggplot(df2, aes(x = genotype, y = mean)) +
  geom_bar(
    stat = "identity",
    fill = palette_geno
  ) +
  geom_errorbar(aes(x = genotype, ymin = mean - sd, ymax = mean + sd),
    width = 0.2,
    linewidth = 0.2
  ) +
  geom_jitter(
    data = df, aes(y = numCells, x = genotype),
    width = 0.2, size = 1
  ) +
  xlab("") +
  ylab("") +
  ggtitle("Nuclei") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 8)
  )

p2 <- ggplot(combined_neuronal@meta.data, aes(x = genotype, y = nFeature_RNA, fill = sample)) +
  geom_boxplot(
    outlier.color = NA,
    linewidth = 0.2,
    varwidth = TRUE
  ) +
  scale_fill_manual(values = c(
    WT_F_1 = "#eecc67",
    WT_F_2 = "#eecc67",
    WT_F_3 = "#eecc67",
    WT_F_4 = "#eecc67",
    WT_F_5 = "#eecc67",
    WT_F_6 = "#eecc67",
    WT_F_7 = "#eecc67",
    WT_F_8 = "#eecc67",
    WT_F_9 = "#eecc67",
    WT_F_10 = "#eecc67",
    Het_F_1 = "#98782d",
    Het_F_2 = "#98782d",
    Het_F_3 = "#98782d",
    Het_F_4 = "#98782d",
    Het_F_5 = "#98782d",
    Het_F_6 = "#98782d",
    WT_M_1 = "#eecc67",
    WT_M_2 = "#eecc67",
    WT_M_3 = "#eecc67",
    WT_M_4 = "#eecc67",
    WT_M_5 = "#eecc67",
    WT_M_6 = "#eecc67",
    WT_M_7 = "#eecc67",
    WT_M_8 = "#eecc67",
    WT_M_9 = "#eecc67",
    WT_M_10 = "#eecc67",
    Het_M_1 = "#98782d",
    Het_M_2 = "#98782d",
    Het_M_3 = "#98782d",
    Het_M_4 = "#98782d",
    Het_M_5 = "#98782d",
    Het_M_6 = "#98782d",
    Het_M_7 = "#98782d",
    Het_M_8 = "#98782d",
    Het_M_9 = "#98782d",
    Het_M_10 = "#98782d"
  )) +
  xlab("") +
  ylab("") +
  ggtitle("Genes") +
  ylim(0, 4000) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 8)
  )

p3 <- ggplot(combined_neuronal@meta.data, aes(x = genotype, y = nCount_RNA, fill = sample)) +
  geom_boxplot(
    outlier.color = NA,
    linewidth = 0.2,
    varwidth = TRUE
  ) +
  scale_fill_manual(values = c(
    WT_F_1 = "#eecc67",
    WT_F_2 = "#eecc67",
    WT_F_3 = "#eecc67",
    WT_F_4 = "#eecc67",
    WT_F_5 = "#eecc67",
    WT_F_6 = "#eecc67",
    WT_F_7 = "#eecc67",
    WT_F_8 = "#eecc67",
    WT_F_9 = "#eecc67",
    WT_F_10 = "#eecc67",
    Het_F_1 = "#98782d",
    Het_F_2 = "#98782d",
    Het_F_3 = "#98782d",
    Het_F_4 = "#98782d",
    Het_F_5 = "#98782d",
    Het_F_6 = "#98782d",
    WT_M_1 = "#eecc67",
    WT_M_2 = "#eecc67",
    WT_M_3 = "#eecc67",
    WT_M_4 = "#eecc67",
    WT_M_5 = "#eecc67",
    WT_M_6 = "#eecc67",
    WT_M_7 = "#eecc67",
    WT_M_8 = "#eecc67",
    WT_M_9 = "#eecc67",
    WT_M_10 = "#eecc67",
    Het_M_1 = "#98782d",
    Het_M_2 = "#98782d",
    Het_M_3 = "#98782d",
    Het_M_4 = "#98782d",
    Het_M_5 = "#98782d",
    Het_M_6 = "#98782d",
    Het_M_7 = "#98782d",
    Het_M_8 = "#98782d",
    Het_M_9 = "#98782d",
    Het_M_10 = "#98782d"
  )) +
  xlab("") +
  ylab("") +
  ggtitle("UMIs") +
  ylim(0, 7500) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 8)
  )

wrap_plots(p1, p2, p3, ncol = 3)

ggsave("../../results/neuronal_processing/final-figs/250320_combined-neuronal_1-5_5pct_genoQC.pdf",
  device = pdf,
  width = 9,
  height = 3
)
```

And a plot of the total number of nuclei per genotype.
```{r}
genotypes_df <- as.data.frame(table(combined_neuronal@meta.data$genotype))
colnames(genotypes_df) <- c("Genotype", "Count")

ggplot(genotypes_df, aes(x = Genotype, y = Count, fill = Genotype)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palette_geno) +
  coord_flip() +
  theme_classic() +
  labs(x = "Genotype", y = "Count")


# Uncomment this to save the figure.
ggsave("../../results/neuronal_processing/final-figs/250320_totalnucspergeno.pdf",
  device = pdf,
  width = 9,
  height = 3
)
```

## for Sex.
```{r add sex to metadata}
meta <- combined_neuronal@meta.data$sample
new_meta <- gsub(".*_([^_]+)_.*", "\\1", levels(meta))
levels(meta) <- new_meta

combined_neuronal <- AddMetaData(combined_neuronal, meta, col.name = "sex")
```

```{r df stats sex}
df <- data.frame(
  numCells = as.vector(samplesTable),
  sex = c(rep(c("F"), times = 6), rep(c("M"), times = 10), rep(c("F"), times = 10), rep(c("M"), times = 10))
)

df$sex <- factor(df$sex, levels = c("M", "F"))

averages <- aggregate(numCells ~ sex, data = df, FUN = mean)
sds <- aggregate(numCells ~ sex, data = df, FUN = function(x) sd(x))

df2 <- merge(averages, sds, by = "sex")
colnames(df2) <- c("sex", "mean", "sd")

df2$sex <- factor(df2$sex, levels = c("M", "F"))
df2 <- df2[order(df2$sex), ]

df
df2
```

```{r bar and boxplots sex}
p1 <- ggplot(df2, aes(x = sex, y = mean)) +
  geom_bar(
    stat = "identity",
    fill = palette_sex
  ) +
  geom_errorbar(aes(x = sex, ymin = mean - sd, ymax = mean + sd),
    width = 0.2,
    linewidth = 0.2
  ) +
  geom_jitter(
    data = df, aes(y = numCells, x = sex),
    width = 0.2, size = 1
  ) +
  xlab("") +
  ylab("") +
  ggtitle("Nuclei") +
  theme_classic() +
  theme(
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 8)
  )

p2 <- ggplot(combined_neuronal@meta.data, aes(x = sex, y = nFeature_RNA, fill = sample)) +
  geom_boxplot(
    outlier.color = NA,
    linewidth = 0.2,
    varwidth = TRUE
  ) +
  scale_fill_manual(values = c(
    WT_F_1 = "#ED6677",
    WT_F_2 = "#ED6677",
    WT_F_3 = "#ED6677",
    WT_F_4 = "#ED6677",
    WT_F_5 = "#ED6677",
    WT_F_6 = "#ED6677",
    WT_F_7 = "#ED6677",
    WT_F_8 = "#ED6677",
    WT_F_9 = "#ED6677",
    WT_F_10 = "#ED6677",
    Het_F_1 = "#ED6677",
    Het_F_2 = "#ED6677",
    Het_F_3 = "#ED6677",
    Het_F_4 = "#ED6677",
    Het_F_5 = "#ED6677",
    Het_F_6 = "#ED6677",
    WT_M_1 = "#4478AB",
    WT_M_2 = "#4478AB",
    WT_M_3 = "#4478AB",
    WT_M_4 = "#4478AB",
    WT_M_5 = "#4478AB",
    WT_M_6 = "#4478AB",
    WT_M_7 = "#4478AB",
    WT_M_8 = "#4478AB",
    WT_M_9 = "#4478AB",
    WT_M_10 = "#4478AB",
    Het_M_1 = "#4478AB",
    Het_M_2 = "#4478AB",
    Het_M_3 = "#4478AB",
    Het_M_4 = "#4478AB",
    Het_M_5 = "#4478AB",
    Het_M_6 = "#4478AB",
    Het_M_7 = "#4478AB",
    Het_M_8 = "#4478AB",
    Het_M_9 = "#4478AB",
    Het_M_10 = "#4478AB"
  )) +
  xlab("") +
  ylab("") +
  ggtitle("Genes") +
  ylim(0, 4000) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 8)
  )


p3 <- ggplot(combined_neuronal@meta.data, aes(x = sex, y = nCount_RNA, fill = sample)) +
  geom_boxplot(
    outlier.color = NA,
    linewidth = 0.2,
    varwidth = TRUE
  ) +
  scale_fill_manual(values = c(
    WT_F_1 = "#ED6677",
    WT_F_2 = "#ED6677",
    WT_F_3 = "#ED6677",
    WT_F_4 = "#ED6677",
    WT_F_5 = "#ED6677",
    WT_F_6 = "#ED6677",
    WT_F_7 = "#ED6677",
    WT_F_8 = "#ED6677",
    WT_F_9 = "#ED6677",
    WT_F_10 = "#ED6677",
    Het_F_1 = "#ED6677",
    Het_F_2 = "#ED6677",
    Het_F_3 = "#ED6677",
    Het_F_4 = "#ED6677",
    Het_F_5 = "#ED6677",
    Het_F_6 = "#ED6677",
    WT_M_1 = "#4478AB",
    WT_M_2 = "#4478AB",
    WT_M_3 = "#4478AB",
    WT_M_4 = "#4478AB",
    WT_M_5 = "#4478AB",
    WT_M_6 = "#4478AB",
    WT_M_7 = "#4478AB",
    WT_M_8 = "#4478AB",
    WT_M_9 = "#4478AB",
    WT_M_10 = "#4478AB",
    Het_M_1 = "#4478AB",
    Het_M_2 = "#4478AB",
    Het_M_3 = "#4478AB",
    Het_M_4 = "#4478AB",
    Het_M_5 = "#4478AB",
    Het_M_6 = "#4478AB",
    Het_M_7 = "#4478AB",
    Het_M_8 = "#4478AB",
    Het_M_9 = "#4478AB",
    Het_M_10 = "#4478AB"
  )) +
  xlab("") +
  ylab("") +
  ggtitle("UMIs") +
  ylim(0, 7500) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14),
    axis.text.y = element_text(size = 8)
  )

wrap_plots(p1, p2, p3, ncol = 3)

# Uncomment this to save the figure.
ggsave("../../results/neuronal_processing/final-figs/250320_combined-neuronal_1-5_5pct_sexQC.pdf",
  device = pdf,
  width = 9,
  height = 3
)
```

And a plot of the total number of nuclei per sex
```{r}
sex_df <- as.data.frame(table(combined_neuronal@meta.data$sex))
colnames(sex_df) <- c("Sex", "Count")

ggplot(sex_df, aes(x = Sex, y = Count, fill = Sex)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = palette_sex) +
  coord_flip() +
  theme_classic() +
  labs(x = "Sex", y = "Count")

# Uncomment this to save the figure.
ggsave("../../results/neuronal_processing/final-figs/250320_totalnucspersex.pdf",
  device = pdf,
  width = 9,
  height = 3
)
```

```{r save}
saveRDS(combined_neuronal, "../../data/250320_combined_neuronal_subsetted.rds")
```
