---
title: "Analysis of P1 MYT1L snRNAseq"
author: "Simona Sarafinovska"
date: "2024-06-01"
---

# Set up environment
```{r dependencies, warning=FALSE, message=FALSE}
# Change this to your directory
.libPaths("/ref/jdlab/software/r-envs/simona/r-packages/4.2/")

# Load dependencies
library(dplyr)
library(Seurat)
library(scCustomize)
library(ggplot2)
library(patchwork)
library(plotly)
library(clustree)

set.seed(420)

source("../color_utils.R")
```

# Load in preprocessed data
```{r load}
combined_neuronal_subsetted <- readRDS("../../data/250320_combined_neuronal_subsetted.rds")
combined_neuronal_subsetted
```
You should have an object with 127456 nuclei and 37462 features.

# Normalize the data

The combined.neuronal data is re-scaled using a global-scaling normalization method. First we normalize the feature expression measurements for each cell by the total expression of the feature, scale using a factor of `10,000`, then the result is log-transformed.

```{r normalize}
combined_neuronal_subsetted <- NormalizeData(combined_neuronal_subsetted,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)
```

# Identification of highly variable features

```{r ID variable features}
combined_neuronal_subsetted <- FindVariableFeatures(combined_neuronal_subsetted,
  selection.method = "vst",
  nfeatures = 3000
)
```

```{r plot variable features}
VariableFeaturePlot_scCustom(seurat_object = combined_neuronal_subsetted, num_features = 20, repel = TRUE)
```

# Scale the data

We apply a linear transformation of the data so that highly-expressed genes do not dominate the analysis. First the expression of each gene is shifted so the mean expression across all cells is 0. Then it scales the expression so the variance across cells is 1.

```{r scale}
all.genes <- rownames(combined_neuronal_subsetted)
combined_neuronal_subsetted <- ScaleData(combined_neuronal_subsetted, features = all.genes)
rm(all.genes)
```

# Dimensional reduction

We perform PCA on the scaled data using the most variable features determined above.

```{r run PCA}
combined_neuronal_subsetted <- RunPCA(combined_neuronal_subsetted,
  npcs = 100,
  features = VariableFeatures(object = combined_neuronal_subsetted)
)
```

```{r VizDim PCA plot}
VizDimLoadings(combined_neuronal_subsetted, dims = 1:4, reduction = "pca") &
  theme(
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8, face = "bold")
  )
```

```{r PCA dim heatmap plot}
DimHeatmap(combined_neuronal_subsetted, dims = 1:12, cells = 1000, balanced = TRUE)
```

First we'll make some elbow plots that plot the standard deviations of the PCs. Look for an "elbow" in the plot where the variance explained by each subsequent PC drops off. The number of PCs at this point is often a good choice

```{r PC elbow plot 100 dims}
ElbowPlot(combined_neuronal_subsetted, ndims = 100)
```

50 PCs appear to be appropriate, so we cluster with 50 dimensions.
```{r find neighbors}
combined_neuronal_subsetted <- FindNeighbors(combined_neuronal_subsetted, dims = 1:50)
```

## Consider resolutions.
```{r find clusters}
combined_neuronal_subsetted <- FindClusters(combined_neuronal_subsetted,
  resolution = c(0.01, 0.05, 0.1, 0.5, 1, 1.2, 1.4, 1.6, 1.8, 2)
)
```


```{r, fig.height=8}
clustree(combined_neuronal_subsetted, prefix = "RNA_snn_res.", layout = "sugiyama")

ggsave(
  filename = "../../results/neuronal_processing/intermed-figs/250321_clustertree.pdf",
  width = 8,
  height = 16,
  dpi = 300
)
```

We'll chose the clustering resolution to be `1.8` so we'll set that as the active identity.
```{r choose resolution}
Idents(combined_neuronal_subsetted) <- combined_neuronal_subsetted$RNA_snn_res.1.8
```

```{r UMAP}
combined_neuronal_subsetted <- RunUMAP(combined_neuronal_subsetted, dims = 1:50, n.components = 2L)
```

# Plot UMAPs
```{r plot UMAP}
DimPlot_scCustom(combined_neuronal_subsetted,
  colors_use = palette_90,
  reduction = "umap",
  label = TRUE,
  raster = FALSE,
  split_seurat = TRUE,
  pt.size = 0.01
) +
  theme(legend.position = "none")

ggsave(
  filename = "../../results/neuronal_processing/final-figs/250321_umap_combined_neuronal_res1.8.pdf",
  width = 12,
  height = 12,
  dpi = 300
)
```

```{r plot UMAP by sample}
DimPlot_scCustom(combined_neuronal_subsetted,
  group.by = "sample",
  reduction = "umap",
  label = FALSE,
  raster = FALSE,
  split_seurat = TRUE,
  pt.size = 0.01
) +
  theme(legend.position = "none")

ggsave(
  filename = "../../results/neuronal_processing/intermed-figs/250321_umap_combined_neuronal_res1.8_bysample.pdf",
  width = 12,
  height = 12,
  dpi = 300
)
```

This will create a series of UMAPs where each cluster is highlighted, which is helpful in identifying if clustering parameters need to be fine tuned. Highlighted clusters should be close together and not spread across UMAP space.
```{r plot highlighting clusters}
dir.create("../../results/neuronal_processing/intermed-figs/IterateCluster_combined_neuronal_res1.8_50dims_250321")

Iterate_Cluster_Highlight_Plot(combined_neuronal_subsetted,
  highlight_color = "navy",
  background_color = "lightgrey",
  file_path = "../../results/neuronal_processing/intermed-figs/IterateCluster_combined_neuronal_res1.8_50dims_250321/",
  file_name = "iterate",
  file_type = ".png",
  raster = FALSE,
  dpi = 100
)
```

We can see how many cells we have for each cluster in a table format.
```{r table of cells per cluster}
table(combined_neuronal_subsetted@meta.data$RNA_snn_res.1.8)
```

# Save!
```{r save}
saveRDS(combined_neuronal_subsetted, "../../data/250321_combined_neuronal_clustered.rds")
```
