---
title: "Analysis of P1 MYT1L snRNAseq"
author: "Simona Sarafinovska"
date: "2024-06-01"
---

# Set up environment
```{r dependencies, warning=FALSE, message=FALSE}
# Change this to your directory
.libPaths('/ref/jdlab/software/r-envs/simona/r-packages/4.2/')

# Load dependencies
library(dplyr)
library(Seurat)
library(scCustomize)
library(ggplot2)
library(patchwork)
library(plotly)
library(clustree)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(anndata)

set.seed(420)

source("../color_utils.R")
```


# Load in preprocessed data
```{r load}
combined_neuronal_clustered <- readRDS("../../data/250321_combined_neuronal_clustered.rds")
combined_neuronal_clustered 
```
You should have a Seurat object with 37462 features across 127456 nuclei.

# Finding cluster markers

```{r findMarkers, message=FALSE, results=FALSE}
combined_neuronal_clustered_markers_all <- FindAllMarkers(combined_neuronal_clustered,
                                             only.pos = FALSE,
                                             min.pct = 0.25,
                                             logfc.threshold = 0.5,
                                             return.thresh = 0.01)

write.csv(combined_neuronal_clustered_markers_all,file = "../../results/neuronal_processing/intermed-figs/250321_markers_neuronal_clustered_res1.8_50dims.csv")

saveRDS(combined_neuronal_clustered_markers_all,file = "../../results/neuronal_processing/intermed-figs/250321_markers_neuronal_clustered_res1.8_50dims.rds")
```


Plotting:
```{r marker heatmap 1}
combined_neuronal_clustered_markers_all %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
plot <- DoHeatmap(combined_neuronal_clustered, features = top10$gene) + NoLegend()
plot

 ggsave(filename = "../../results/neuronal_processing/intermed-figs/250321_markers_neuronal_clustered_res1.8_50dims.pdf",
         plot = plot,
         width = 18,
         height = 18,
         dpi = 300)
```

```{r marker heatmap 2}
 combined_neuronal_clustered_markers_all %>%
   group_by(cluster) %>%
   top_n(n = 10, wt = avg_log2FC) -> top10

 plot <- DoHeatmap(combined_neuronal_clustered, 
                   feature = top10$gene,
                   #group.colors = palette,
                   label = FALSE,
                   disp.min = 0,
                   disp.max = 2,
                   raster = TRUE) + 
   scale_fill_viridis() +
   NoLegend()
 plot
  ggsave(filename = "../../results/neuronal_processing/intermed-figs/250321_markers_neuronal_clustered_res1.8_50dims-2.pdf",
         plot = plot,
         width = 18,
         height = 24,
         dpi = 300)
```

```{r filtered heatmap}
# Filtered list of neuronal genes by functional category
neuronal.genes <- c(
  # Neurotransmitter systems
  "Vipr2", "Hcrtr2", "Grm8", "Grm3", "Slc18a2", "Slc6a3", "Drd2", "Chrm3", 
  "Htr2c", "Tacr3", "Gria1", "Grik1", "Gad2", "Grik3", "Htr1f", "Chrm2", 
  "Tacr1", "Slc5a7", "Chat", "Gabra4", "Grm1",
  
  # Ion channels
  "Kcnh1", "Kcnj6", "Kcnma1", "Kcnh8", "Kcnab1", "Cacna2d2", "Asic2", "Kcnq5", 
  "Trpm3", "Cacna2d3", "Kcnmb2", "Hcn1", "Kcnip1", "Scn9a", "Kcnb2", "Cacna2d1", 
  "Kcnh5", "Kcnc2", "Trpm6", "Kcnip4",
  
  # Synaptic function
  "Syt9", "Stxbp6", "Sv2b", "Unc13c", "Nrxn3", "Synpr", "Baiap3", "Nptxr", 
  "Dlgap2", "Rimbp2", "Cadps2",
  
  # Axon guidance and cell adhesion
  "Cdh13", "Nrg1", "Tenm3", "Sema5a", "Slit2", "Robo2", "Sema6d", "Slit3", 
  "Sema3a", "Cntnap5a", "Lrrc4c", "Cdh23", "Ntm", "Cdh18", "Cdh12", "Cdh8", 
  "Sdk1", "Cntn5", "Pcdh7", "Pcdh15", "Ntng2", "Cntn6", "Kirrel3", "Cntnap5c", "Ntng1",
  
  # Transcription factors and development
  "Nfia", "Ebf2", "Pbx1", "Mef2c", "Foxp1", "Lhx8", "Pax6", "Ebf3", "Ebf1", 
  "Sim1", "Lmx1a", "Zfhx4", "Zfhx3", "Pou6f2", "Zfpm2", "Onecut2", "Meis1", 
  "Nfib", "Tcf4", "Tshz1", "Meis2", "Foxp2", "Pbx3", "Satb2", "Tcf7l2", "Sox6", "Sox5"
)

# Filter the combined_neuronal_clustered_markers_all dataset
neuronal.markers.filtered <- combined_neuronal_clustered_markers_all %>%
  filter(gene %in% neuronal.genes)

# Get top markers per cluster with higher specificity
neuronal.top.markers <- neuronal.markers.filtered %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC) # Reduced from 10 to 5 genes per cluster

# Create the heatmap with better visibility
plot <- DoHeatmap(combined_neuronal_clustered, 
          features = neuronal.top.markers$gene,
          label = TRUE, # Now we can show labels as we have fewer genes
          disp.min = 0,
          disp.max = 2,
          raster = TRUE,
          size = 3) + # Increase text size
  scale_fill_viridis() +
  theme(axis.text.y = element_text(size = 8)) +
  NoLegend()

# Preview the plot
plot

ggsave(filename = "../../results/neuronal_processing/intermed-figs/250321_selectedmarkers_neuronal_clustered_res1.8_50dims.pdf",
         plot = plot,
         width = 18,
         height = 24,
          dpi = 300)
```

Sorting by region:
```{r heatmap by region, fig.height=10}
# Get average expression matrix for the top marker genes
avg_expr <- AverageExpression(combined_neuronal_clustered, 
                             features = unique(neuronal.top.markers$gene),
                             group.by = "ident",
                             assays = "RNA", 
                             slot = "scale.data") 

# Convert to matrix
expr_matrix <- as.matrix(avg_expr$RNA)

# Define the regional order
region_order <- c("Amygdala/Hypothalamus", "Hypothalamus", "Midbrain", 
                 "Striatum/Pallidum", "Thalamus", "Zona Incerta")

# Get the cluster names and their corresponding regions
clusters <- colnames(expr_matrix)
cluster_regions <- combined_neuronal_clustered$anatomical_region[match(clusters, Idents(combined_neuronal_clustered))]

# Create column annotation for pheatmap with both Region and Cluster
anno_col <- data.frame(
  Region = cluster_regions,
  Cluster = clusters,  # Add the cluster IDs as a separate annotation
  row.names = clusters
)

# Convert region to factor with proper order
anno_col$Region <- factor(anno_col$Region, levels = region_order)


# Use your palette_70 for cluster colors
# Assuming palette_70 is a vector with at least as many colors as clusters
cluster_colors <- palette_70[1:length(clusters)]
names(cluster_colors) <- clusters

# Create the annotation_colors list with both annotations
anno_colors <- list(
  Region = region_muted_colors,
  Cluster = cluster_colors
)

# Sort columns by region
anno_col_sorted <- anno_col[order(anno_col$Region),, drop = FALSE]
expr_matrix_sorted <- expr_matrix[, rownames(anno_col_sorted)]

# Create the heatmap with both annotations
p <- pheatmap(
  expr_matrix_sorted,
  annotation_col = anno_col_sorted,
  annotation_colors = anno_colors,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_colnames = FALSE,
  fontsize_row = 8,
  fontsize_col = 8,
  color = viridisLite::viridis(100),
  annotation_legend = FALSE,
  breaks = seq(0, 2, length.out = 100),  
  legend_breaks = seq(0, 2, 0.5),        
  legend_labels = c("0", "0.5", "1", "1.5", "2")
)

save_complexheatmap_pdf(
  p, 
  "../../results/neuronal_processing/final_figs/250321_neuronal_markers_by_region.pdf", 
  width = 12, 
  height = 14
)
```


# Hypomap
We followed the vignette here: https://github.com/lsteuernagel/mapscvi to  map combined onto HypoMap (Steuernagel et al., Nature Met., 2022) and obtain predicted annotations. 

```{r import hypomap annotations}
hypomapped_neuronal <- readRDS("../../data/240602_hypomapped_neuronal_1-5_5pct.rds")
hypomapped_neuronal
```
You should have the same features and nuclei (37462 features across 127456 nuclei), but more dimensional reductions (now have scvi and umap_scvi).

```{r add annotations to combined}
cols_to_transfer <- c("prediction_probability", "predicted", "avg_neighbor_distance")

for (col in cols_to_transfer) {
  new_col_name <- paste0("hypomap_", col)
  
  combined_neuronal_clustered[[new_col_name]] <-
    hypomapped_neuronal@meta.data[[col]][match(colnames(combined_neuronal_clustered),
                                      colnames(hypomapped_neuronal))]
}
```


# MapMyCells

## Create minimal h5ad
```{r create h5ad}
ph <- mapIds(org.Mm.eg.db,
             keys = rownames(combined_neuronal_clustered),
             column = "ENSEMBL",
             keytype = "SYMBOL",
             multiVals = "first")
ph <- as.data.frame(ph)

df <- data.frame(
  GeneSymbol = rownames(combined_neuronal_clustered),
  EnsemblID = ph
)
df <- na.omit(df)
duplicates <- df$ph[duplicated(df$ph) | duplicated(df$ph, fromLast = TRUE)]
df_dups_removed <- df %>% distinct(ph, .keep_all = TRUE)

genes_to_remove <- rownames(combined_neuronal_clustered)[is.na(ph)]
combined_neuronal_clustered_cleaned <- combined_neuronal_clustered[rownames(combined_neuronal_clustered) %in% df_dups_removed$GeneSymbol, ]

ph <- mapIds(org.Mm.eg.db,
             keys = rownames(combined_neuronal_clustered_cleaned),
             column = "ENSEMBL",
             keytype = "SYMBOL",
             multiVals = "first")
ph <- as.data.frame(ph)

df <- data.frame(
  GeneSymbol = rownames(combined_neuronal_clustered_cleaned),
  EnsemblID = ph
)

raw_counts <- as.matrix(combined_neuronal_clustered_cleaned@assays$RNA@counts)
raw_counts <- GetAssayData(object = combined_neuronal_clustered_cleaned,
                           slot = "counts")
colnames(raw_counts) <- rownames(combined_neuronal_clustered_cleaned@meta.data)
samples <- colnames(raw_counts)
genes <- df$ph
raw_counts <- t(raw_counts)

adata <- AnnData(X = raw_counts,
                 var = data.frame(genes = genes, row.names = genes),
                 obs = data.frame(samples = samples, row.names = samples))


write_h5ad(adata, "../../data/250321_combined_neuronal_clustered_1-5_5pct_res1.8.h5ad")
remove(combined_neuronal_clustered_cleaned)
```
And you can take the .h5ad file and upload it to MapMyCells on the Allen Brain Cell Atlas site!

## Bring in MapMyCells Data
Now we load in the csv provided by MapMyCells to see what it predicts for our cluster annotations. Make sure to delete the first four rows before uploading the .csv to your folder.
```{r read in MapMyCells data}
ABC_map <- read.csv("../../data/250321_combined_neuronal_clustered_1-5_5pct_res18_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1742946742715.csv")


combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$class_name,col.name='ABC_class_name')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$class_bootstrapping_probability,col.name='ABC_class_probability')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$class_label,col.name='ABC_class_label')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$subclass_name,col.name='ABC_subclass_name')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$subclass_bootstrapping_probability,col.name='ABC_subclass_probability')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$subclass_label,col.name='ABC_subclass_label')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$supertype_name,col.name='ABC_supertype_name')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$supertype_bootstrapping_probability,col.name='ABC_supertype_probability')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$supertype_label,col.name='ABC_supertype_label')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$cluster_name,col.name='ABC_cluster_name')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$cluster_bootstrapping_probability,col.name='ABC_cluster_probability')
combined_neuronal_clustered <- AddMetaData(object = combined_neuronal_clustered,metadata=ABC_map$cluster_label,col.name='ABC_cluster_label')
```

This will print out a csv with Seurat clusters on the columns and ABC clusters on the rows so you can sort and see how they corelate!
```{r save table with mapmycells clusters}
clusters <- combined_neuronal_clustered@meta.data$RNA_snn_res.1.8
ABC_class_names <- combined_neuronal_clustered@meta.data$ABC_class_name
ABC_subclass_names <- combined_neuronal_clustered@meta.data$ABC_subclass_name
ABC_supertype_name <- combined_neuronal_clustered@meta.data$ABC_supertype_name
ABC_cluster_name <- combined_neuronal_clustered@meta.data$ABC_cluster_name

df <- data.frame(clusters, ABC_class_names, ABC_subclass_names,ABC_supertype_name, ABC_cluster_name)

table <- table(df$ABC_class_names, df$clusters)
write.csv (table, "../../results/neuronal_processing/250321_clustersbyABCclass_neuronal.csv", row.names = TRUE)

table <- table(df$ABC_subclass_names, df$clusters)
write.csv (table, "../../results/neuronal_processing/250321_clustersbyABCsubclass_neuronal.csv", row.names = TRUE)

table <- table(df$ABC_supertype_name, df$clusters)
write.csv (table, "../../results/neuronal_processing/250321_clustersbyABCsupertype_neuronal.csv", row.names = TRUE)

table <- table(df$ABC_cluster_name, df$clusters)
write.csv (table, "../../results/neuronal_processing/250321_clustersbyABCcluster_neuronal.csv", row.names = TRUE)
```

Plot a histogram of how many cells are in each ABC class per cluster.
```{r ABC class histogram}
table_df <- data.frame(table(df$ABC_class_names, df$clusters))
colnames(table_df) <- c("Class_Name", "Seurat_Cluster","Number_of_Cells")

clusters <- seq(0,max(as.numeric(df$clusters))-1)

dir.create("../../results/neuronal_processing/intermed-figs/ClassHistograms_250321")

for (i in clusters) {
  p <- table_df %>%
    #use filter to subset the rows based on the Var2 column values
    filter(Seurat_Cluster == i) %>% 
    filter(Number_of_Cells > 0) %>%
    ggplot(aes(x=Class_Name, y=Number_of_Cells)) +
    geom_bar(stat="identity") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
  file_name = paste("../../results/neuronal_processing/intermed-figs/ClassHistograms_250321/", i, ".png", sep="")
  
  ggsave(filename = file_name,
       width = 12,
       height = 12,
       dpi = 300)
}
```

```{r ABC subclass histogram}
table_df <- data.frame(table(df$ABC_subclass_names, df$clusters))
colnames(table_df) <- c("Subclass_Name", "Seurat_Cluster","Number_of_Cells")

clusters <- seq(0,max(as.numeric(df$clusters))-1)


dir.create("../../results/neuronal_processing/intermed-figs/SubclassHistograms_250321")

for (i in clusters) {
  p <- table_df %>%
    #use filter to subset the rows based on the Var2 column values
    filter(Seurat_Cluster == i) %>% 
    filter(Number_of_Cells > 20) %>%
    ggplot(aes(x=Subclass_Name, y=Number_of_Cells)) +
    geom_bar(stat="identity") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
  file_name = paste("../../results/neuronal_processing/intermed-figs/SubclassHistograms_250321/", i, ".png", sep="")
  
  ggsave(filename = file_name,
       width = 12,
       height = 12,
       dpi = 300)
}
```



# Cluster Tree
We noticed some clusters have significantly overlapping annotations based on MapMyCells, HypoMap, and markers. So we will make a hierarchical cluster tree to identify whether to merge: 
```{r hierarchical cluster tree, fig.height=8}
Idents(combined_neuronal_clustered) <- combined_neuronal_clustered$RNA_snn_res.1.8

combined_neuronal_clustered <- BuildClusterTree(combined_neuronal_clustered, dims = 1:50, reorder = TRUE)

PlotClusterTree(combined_neuronal_clustered)

file_path= "../../results/neuronal_processing/intermed-figs/250321_clustrerhierarchy_neuronal_1-5_5pct_res1.8_50dims.png"
png(height=200, width= 800, file=file_path)

PlotClusterTree(combined_neuronal_clustered)

dev.off()
```

Based on manual comparison of clusters at each terminal node and leaf, we have dedcided to merge clusters as below:
```{r merging clusters}
merged_idents <- c("66",
  "8",
  "17",
  "17",
  "17",
  "17",
  "39",
  "43",
  "69",
  "18",
  "4",
  "23",
  "58",
  "1",
  "71",
  "61",
  "55",
  "21",
  "21",
  "25",
  "45",
  "0",
  "63",
  "22",
  "52",
  "36",
  "35",
  "31",
  "42",
  "12",
  "46",
  "38",
  "5",
  "28",
  "19",
  "3",
  "37",
  "10",
  "6",
  "72",
  "32",
  "11",
  "7",
  "57",
  "62",
  "14",
  "67",
  "74",
  "2",
  "20",
  "49",
  "26",
  "44",
  "40",
  "15",
  "15",
  "13",
  "13",
  "29",
  "33",
  "65",
  "34",
  "34",
  "56",
  "48",
  "75",
  "9",
  "9",
  "47",
  "59",
  "53",
  "60",
  "24",
  "51",
  "50",
  "50")
names(merged_idents) <- levels(combined_neuronal_clustered)
merged_idents
combined_neuronal_clustered <- RenameIdents(combined_neuronal_clustered, merged_idents)
combined_neuronal_clustered@meta.data$merged_idents <- Idents(combined_neuronal_clustered)
```

# Assigning idents
Based on the annotation tools above, we have manually annotated the clusters as below:
```{r adding cell_type idents}
Idents(combined_neuronal_clustered) <- combined_neuronal_clustered$merged_idents

cell_type <- c(
  "66" = "TH Prkcd Grin2c Glut_1",
  "8" = "TH Prkcd Grin2c Glut_2",
  "17" = "TH Prkcd Grin2c Glut_3",
  "39" = "TH Prkcd Grin2c Glut_4",
  "43" = "AV Col27a1 Glut",
  "69" = "TH Prkcd Grin2c Glut_5",
  "18" = "TH Prkcd Grin2c Glut_6",
  "4" = "TH Prkcd Grin2c Glut_7",
  "23" = "RE-Xi Nox4 Glut",
  "58" = "TH-PVT-RE Ntrk1-Nox4 Glut",
  "1" = "TH Sema5b-Ntrk Glut",
  "71" = "AD Serpinb7 Glut",
  "61" = "HY NSCs_1",
  "55" = "Chat GABA",
  "21" = "STR-PAL Chst9 Gaba",
  "25" = "Microglia",
  "45" = "OPC",
  "0" = "Mixed",
  "63" = "VMH Nr5a1 Glut_1",
  "22" = "Lhx6/Lhx8 Gaba",
  "52" = "Lhx6 Gaba",
  "36" = "PH-LHA-MM Foxb1 Glut",
  "35" = "PV-PT Ntrk1/Sp9 Glut",
  "31" = "PO-MEA Glut",
  "42" = "PO-COAa-PAA-MEA Barhl2 Glut",
  "12" = "VMH Nr5a1 Glut_2",
  "46" = "DMH Hmx2/Vgll1 Glut",
  "38" = "TU-ARH Otp Six6 Gaba",
  "5" = "DMH Gaba",
  "28" = "ZI/PVH/SBPV Pax6/Prox1 Gaba",
  "19" = "AHN Onecut3 Gaba",
  "3" = "AHN-PVHd-PVR Mixed Gaba",
  "37" = "PVpo-VMPO-MPN Hmx2 Gaba",
  "10" = "PO-CEA-BST Meis2 Gaba",
  "6" = "PO-MEA-BST Mixed Gaba",
  "72" = "LHA Pmch Glut",
  "32" = "Pitx2 Glut",
  "11" = "Trh Glut",
  "7" = "PH-LHA-PVH-SO-PVa Nxph4 Glut",
  "57" = "RT-ZI Gnb3 Gaba_1",
  "62" = "GPi Tbr1 Cngb3 Gaba-Glut",
  "14" = "SI-MA-LPO-LHA Skor1 Glut",
  "67" = "ADP-MPO Trp73 Glut",
  "74" = "SCsg Gabrr2 Gaba",
  "2" = "ZI Mixed Gaba",
  "20" = "ZI Pax6 Gaba",
  "49" = "GPe-SI Sox6 Cyp26b1 Gaba",
  "26" = "NDB-SI-MA-STRv Lhx8 Gaba",
  "44" = "SI-MPO-LPO-ADP Lhx8 Gaba",
  "40" = "HY NSCs_2",
  "15" = "STR D1/D2 Gaba",
  "13" = "STR D1 Gaba",
  "29" = "PO-MEA Slc17a7 Glut",
  "33" = "ARC Mixed Glut",
  "65" = "PVN Mixed Glut",
  "34" = "PF Fzd5 Glut",
  "56" = "LH-MH Glut",
  "48" = "STN-PSTN Pitx2 Glut",
  "75" = "SNc-VTA-RAmb Foxa1 Dopa",
  "9" = "RT-ZI Gnb3 Gaba_2",
  "47" = "PVH-SO-PVa Avp Glut",
  "59" = "PVH-SO-PVa Agtr1a Glut",
  "53" = "TRS-BAC Sln Glut",
  "60" = "Hist Gaba",
  "24" = "Mixed Meis2 Gaba",
  "51" = "TH Prkcd Grin2c Glut_8",
  "50" = "SCH Six6 Cdc14a Gaba"
)
combined_neuronal_clustered <- RenameIdents(combined_neuronal_clustered, cell_type)
combined_neuronal_clustered@meta.data$cell_type <- Idents(combined_neuronal_clustered)
```

##Taking out some more glial clusters

```{r removing non-neuronal clusters}

Idents(combined_neuronal_clustered) <- combined_neuronal_clustered$cell_type

non_neuronal <- c("OPC", 
                  "Microglia") 

combined_neuronal_clustered <- subset(combined_neuronal_clustered,
                             idents = non_neuronal,
                             invert = TRUE)
combined_neuronal_clustered
```
You should now have a Seurat object with 37642 features and 124601 nuclei.

# Adding in behavior data
We need to add mean values from behavior into the single nucleus dataset:
```{r read in behavior data}
behavior_data <- read.csv("../../data/073123_SocialOperant_short.csv", stringsAsFactors = FALSE)
behavior_data <- behavior_data[,apply(behavior_data, 2, function(x) { sum(!is.na(x)) > 0 })]
head(behavior_data)
```

```{r make new metadata file}
combined_neuronal_clustered@meta.data$sample <- as.factor(gsub("-", "_", combined_neuronal_clustered@meta.data$sample))

# Merge the metadata with behavior_data
metadata <- merge(combined_neuronal_clustered@meta.data, behavior_data, by = "sample", all.x = TRUE, sort = FALSE)

# Set the rownames of the merged metadata to the preserved rownames
rownames(metadata) <- rownames(combined_neuronal_clustered@meta.data) 

dim(metadata)
dim(combined_neuronal_clustered@meta.data)
```

```{r merge metadata}
combined_neuronal_clustered@meta.data <- metadata
```

```{r rename values}
# Rename the values in the Sac_Learner column
combined_neuronal_clustered$Sac_Learner <- ifelse(combined_neuronal_clustered$Sac_Learner == 1, "Learner", "Non-Learner")

# Convert the Sac_Learner column to a factor
combined_neuronal_clustered$Sac_Learner <- factor(combined_neuronal_clustered$Sac_Learner)

# Rename the levels of the factor
levels(combined_neuronal_clustered$Sac_Learner) <- c("Learner", "Non_Learner")
```

Data on the day of sacrifice was pulled down separately, so we are merging that in here. 
```{r read in behavior data sac}
sac_behavior_data <- read.csv("../../data/073123_SacDay_forR.csv", stringsAsFactors = FALSE)
sac_behavior_data <- sac_behavior_data[,apply(sac_behavior_data, 2, function(x) { sum(!is.na(x)) > 0 })]
head(sac_behavior_data)
```


```{r make new metadata file sac}
combined_neuronal_clustered@meta.data$sample <- as.factor(gsub("-", "_", combined_neuronal_clustered@meta.data$sample))

# Merge the metadata with behavior_data
sac_metadata <- merge(combined_neuronal_clustered@meta.data, sac_behavior_data, by = "sample", all.x = TRUE, sort = FALSE)

# Set the rownames of the merged metadata to the preserved rownames
rownames(sac_metadata) <- rownames(combined_neuronal_clustered@meta.data) 

dim(sac_metadata)
dim(combined_neuronal_clustered@meta.data)
```
```{r merge metadata sac}
combined_neuronal_clustered@meta.data <- sac_metadata
```


# Save!
```{r save}
saveRDS(combined_neuronal_clustered, "../../data/250321_combined_neuronal_annotated.rds")
```
