---
title: "250205_snRNAseq_annotate_clean"
author: "Simona Sarafinovska"
date: "2025-02-05"
---

# Set up environment
We first need to tell RStudio where to find our packages. 
```{r dependencies, warning=FALSE, message=FALSE}
# Change this to your directory
.libPaths('/ref/jdlab/software/r-envs/simona/r-packages/4.2/')

# Load dependencies
library(dplyr)
library(Seurat)
library(scCustomize)
library(ggplot2)
library(patchwork)
library(plotly)
library(clustree)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(anndata)

set.seed(420)

source("../color_utils.R")
```

# Load in Data
```{r load in data}
combined_clustered <- readRDS( "../../data/250221_combined_clustered_1-5_5pct.rds")
combined_clustered
```
You should have an object with 37462 features across 138033 nuclei. 


# Finding cluster markers
```{r findMarkers, message=FALSE, results=FALSE}
combined.markers.all <- FindAllMarkers(combined_clustered,
                                             only.pos = FALSE,
                                             min.pct = 0.25,
                                             logfc.threshold = 0.5,
                                             return.thresh = 0.01)

write.csv(combined.markers.all,file = "../../results/processing/250308_combinedmarkers-5pct_50dims.csv")

saveRDS(combined.markers.all,file = "../../results/processing/250308_combinedmarkers-5pct_50dims.rds")
```

## And plotting heatmaps
```{r}
combined.markers.all %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
plot <- DoHeatmap(combined_clustered, features = top10$gene) + NoLegend()

 ggsave(filename = "../../results/processing/intermed-figs/250308_combinedmarkers_50dim_5pct.pdf",
        plot = plot,
        width = 18,
        height = 18,
        dpi = 300)
```

```{r}
 combined.markers.all %>%
   group_by(cluster) %>%
   top_n(n = 10, wt = avg_log2FC) -> top10_AY

 plot <- DoHeatmap(combined_clustered, 
                   feature = top10_AY$gene,
                   label = FALSE,
                   disp.min = 0,
                   disp.max = 2,
                   raster = TRUE) + 
   scale_fill_viridis() +
   NoLegend()
 plot
 ggsave(filename = "../../results/processing/intermed-figs/250308_combinedmarkers_50dim_5pct-2.pdf",
        plot = plot,
        width = 18,
        height = 24,
        dpi = 300)
```


# Hypomap
We followed the vignette here: https://github.com/lsteuernagel/mapscvi to  map combined onto HypoMap (Steuernagel et al., Nature Met., 2022) and obtain predicted annotations. 

```{r import hypomap annotations}
hypomapped <- readRDS("../../data/240531_hypomapped_1-5_5pct_res2.rds")
hypomapped
```
You should have the same features and nuclei (37462 features across 138033 nuclei), but more dimensional reductions (now have scvi and umap_scvi).

```{r add annotations to combined}
cols_to_transfer <- c("prediction_probability", "predicted", "avg_neighbor_distance")

for (col in cols_to_transfer) {
  new_col_name <- paste0("hypomap_", col)
  
  combined_clustered[[new_col_name]] <-
    hypomapped@meta.data[[col]][match(colnames(combined_clustered),
                                      colnames(hypomapped))]
}
```

# MapMyCells
## Create minimal h5ad
```{r create h5ad}
ph <- mapIds(org.Mm.eg.db,
             keys = rownames(combined_clustered),
             column = "ENSEMBL",
             keytype = "SYMBOL",
             multiVals = "first")
ph <- as.data.frame(ph)

df <- data.frame(
  GeneSymbol = rownames(combined_clustered),
  EnsemblID = ph
)
df <- na.omit(df)
duplicates <- df$ph[duplicated(df$ph) | duplicated(df$ph, fromLast = TRUE)]
df_dups_removed <- df %>% distinct(ph, .keep_all = TRUE)

genes_to_remove <- rownames(combined_clustered)[is.na(ph)]
combined_clustered.cleaned <- combined_clustered[rownames(combined_clustered) %in% df_dups_removed$GeneSymbol, ]

ph <- mapIds(org.Mm.eg.db,
             keys = rownames(combined_clustered.cleaned),
             column = "ENSEMBL",
             keytype = "SYMBOL",
             multiVals = "first")
ph <- as.data.frame(ph)

df <- data.frame(
  GeneSymbol = rownames(combined_clustered.cleaned),
  EnsemblID = ph
)

raw_counts <- GetAssayData(object = combined_clustered.cleaned,
                           slot = "counts")
colnames(raw_counts) <- rownames(combined_clustered.cleaned@meta.data)
samples <- colnames(raw_counts)
genes <- df$ph
raw_counts <- t(raw_counts)

adata <- AnnData(X = raw_counts,
                 var = data.frame(genes = genes, row.names = genes),
                 obs = data.frame(samples = samples, row.names = samples))
write_h5ad(adata, "../../data/250308_combined_clustered_1-5_5pct_res2.h5ad")
remove(combined_clustered.cleaned)
```
And you can take the .h5ad file and upload it to MapMyCells on the Allen Brain Cell Atlas site!

## Bring in MapMyCells Data
Now we load in the csv provided by MapMyCells to see what it predicts for our cluster annotations.
```{r read in MapMyCells data}
ABC_map <- read.csv("../../data/250308_combined_clustered_1-5_5pct_res2_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1741499269497.csv")

combined_clustered <- AddMetaData(object = combined_clustered,metadata=ABC_map$class_name,col.name='ABC_class_name')
combined_clustered <- AddMetaData(object = combined_clustered,metadata=ABC_map$class_bootstrapping_probability,col.name='ABC_class_probability')
combined_clustered <- AddMetaData(object = combined_clustered,metadata=ABC_map$class_label,col.name='ABC_class_label')
combined_clustered <- AddMetaData(object = combined_clustered,metadata=ABC_map$subclass_name,col.name='ABC_subclass_name')
combined_clustered <- AddMetaData(object = combined_clustered,metadata=ABC_map$subclass_bootstrapping_probability,col.name='ABC_subclass_probability')
combined_clustered <- AddMetaData(object = combined_clustered,metadata=ABC_map$subclass_label,col.name='ABC_subclass_label')
```

This will print out a csv with Seurat clusters on the columns and ABC clusters on the rows so you can sort and see how they corelate.
```{r save table with MapMyCells clusters}
clusters <- combined_clustered@meta.data$RNA_snn_res.2
ABC_cluster_names <- combined_clustered@meta.data$ABC_class_name

df <- data.frame(clusters, ABC_cluster_names)

table <- table(df$ABC_cluster_names, df$clusters)
write.csv (table, "../../results/processing/250308_clustersbyABCclass_1-5_5pct_50dims.csv", row.names = TRUE)
```


# Assign Cell Class
Let's use the MapMyCells findings to map names to clusters:
```{r assign cluster names by MapMyCells}
cluster_matrix <- as.matrix(table)

cluster_to_name <- apply(cluster_matrix, 2, function(x) {
    # Get the ABC cluster name with maximum count
    rownames(cluster_matrix)[which.max(x)]
})


new.cluster.ids <- sapply(names(cluster_to_name), function(cluster) {
    abc_name <- cluster_to_name[cluster]
    cell_type <- sub("^\\d+\\s+", "", abc_name)
    paste0(cluster, "-", cell_type)
})

new.cluster.ids
```

```{r rename class idents}
new.cluster.ids <- sort(new.cluster.ids)

names(new.cluster.ids) <- levels(combined_clustered)
combined_clustered <- RenameIdents(combined_clustered, new.cluster.ids)
```

Adding class as a metadata column
```{r add class metadata}
class_assignment <- function(cluster_name) {
  if(grepl("GABA|Glut|IMN|Dopa", cluster_name, ignore.case=TRUE) && 
     !grepl("Astro|OPC|Oligo|Vascular|Immune|OEC", cluster_name, ignore.case=TRUE)) {
    return("Neuronal")
  } else {
    return("Non-Neuronal")
  }
}

combined_clustered$class <- sapply(Idents(combined_clustered), class_assignment)
```

```{r table by class}
class_table <- table(combined_clustered@meta.data$class)
class_table
```
-
```{r umap by class}
DimPlot_scCustom(combined_clustered,
                 group.by = 'class',
                 reduction = 'umap',
                 label = FALSE,
                 raster = FALSE,
                 split_seurat = TRUE,
                 pt.size = 0.01) +
  theme(legend.position = 'right')

ggsave(filename = "../../results/processing/250308_umapbyclass_combined_1-5_5pct_res2.png",
       width = 12,
       height = 12,
       dpi = 300)
```


```{r save}
saveRDS(combined_clustered, "../../data/250308_combined_annotated_1-5_5pct_res2.rds")
```