---
title: "Map My Cells Probability Plot"
author: "Emma Jones"
date: "2025-07-31"
output: html_document
---

The purpose of this script is to generate the visualizations requested by some of our reviewers to clarify cell type assignments. First, we will make a plot of cell probability assignments from map my cells for each cluster, and then will make plots of some of our most notable genes in the paper.

We do not need to independently load in the map my cells results, as cell probability is included in the metadata of our main object.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up environment

```{r dependencies, warning = FALSE, message = FALSE}
# Change this to your package directory
.libPaths("/ref/jdlab/software/r-envs/simona/r-packages/4.2/")

# Load dependencies
library(dplyr)
library(Seurat)
library(patchwork)
library(Matrix)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(scCustomize)
library(patchwork)
library(qs)
library(viridis)
library(ggpubr)
library(SeuratDisk)
library(SeuratWrappers)
library(ComplexHeatmap)
library(stringr)
library(ggrastr)
library(parallel)
library(ggtree)
library(speckle)
library(limma)
library(gt)
library(data.table)
library(biomaRt)
library(tidyr)
library(compositions)
library(lisi)

set.seed(420)

source("../color_utils.R")
```

# Load in preprocessed data

```{r load rds}
combined_neuronal <- readRDS("../../data/250228_combined.neuronal_1-5_5pct.rds")
combined_neuronal
```
# Probability boxplots

First, we can use existing metadata to create probability boxplots.

```{r make bloxplots for probability}
# ggplot for class
class_probability_plot <- 
  ggplot(combined_neuronal@meta.data, aes(y=ABC_class_probability, x=ABC_class_name)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
      labs(x = "Class Names (33)", y = "Class Bootstrapping\nProbability")

class_probability_plot

ggsave("../../results/supplementary/class_probability_plot.png", width = 12, height = 4, dpi = 300)

# ggplot for subclass
subclass_probability_plot <- 
  ggplot(combined_neuronal@meta.data, aes(y=ABC_subclass_probability, x=ABC_subclass_name)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
      labs(x = "Subclass Names (274)", y = "Subclass Bootstrapping\nProbability")

subclass_probability_plot

ggsave("../../results/supplementary/subclass_probability_plot.png", width = 12, height = 4, dpi = 300)

# ggplot for hypomap
hypomap_probability_plot <- 
  ggplot(combined_neuronal@meta.data, aes(y=hypomap_prediction_probability, x=hypomap_predicted)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
      labs(x = "Hypomap Names (166)", y = "Hypomap Prediction\nProbability")

hypomap_probability_plot

ggsave("../../results/supplementary/hypomap_probability_plot.png", width = 12, height = 4, dpi = 300)

```


## Probability plots for class and subclass with at least 30 or 10 cells

```{r make bloxplots for probability}
# Filter combined_neuronal@meta.data to only include classes with at least 30 cells
class_filtered_data <- combined_neuronal@meta.data %>%
  group_by(ABC_class_name) %>%
  filter(n() >= 30) %>%
  ungroup()

# ggplot for class
class_probability_plot <- 
  ggplot(class_filtered_data, aes(y=ABC_class_probability, x=ABC_class_name)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(x = "Class Names (27)", y = "Class Bootstrapping\nProbability")

class_probability_plot

ggsave("../../results/supplementary/class_probability_plot_over30.png", width = 12, height = 4, dpi = 300)

# Filter combined_neuronal@meta.data to only include subclasses with at least 10 cells
subclass_filtered_data <- combined_neuronal@meta.data %>%
  group_by(ABC_subclass_name) %>%
  filter(n() >= 10) %>%
  ungroup()

# ggplot for subclass
subclass_probability_plot <- 
  ggplot(subclass_filtered_data, aes(y=ABC_subclass_probability, x=ABC_subclass_name)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(x = "Subclass Names (203)", y = "Subclass Bootstrapping\nProbability")

subclass_probability_plot

ggsave("../../results/supplementary/subclass_probability_plot_over10.png", width = 12, height = 4, dpi = 300)

# Filter combined_neuronal@meta.data to only include subclasses with at least 10 cells
hypomap_filtered_data <- combined_neuronal@meta.data %>%
  group_by(hypomap_predicted) %>%
  filter(n() >= 20) %>%
  ungroup()

# ggplot for hypomap
hypomap_probability_plot <- 
  ggplot(hypomap_filtered_data, aes(y=hypomap_prediction_probability, x=hypomap_predicted)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
      labs(x = "Hypomap Names (132)", y = "Hypomap Prediction\nProbability")

hypomap_probability_plot

ggsave("../../results/supplementary/hypomap_probability_plot_over20.png", width = 12, height = 4, dpi = 300)

```

# UMAP FeaturePlots of Agtr1a, Nxph4, and Myt1l

```{r make plots for Agtr1a}

agtr1a_umap_overall <- FeaturePlot(combined_neuronal, features = "Agtr1a", raster = FALSE)
agtr1a_umap_overall
ggsave("../../results/supplementary/agtr1a_umap_full.png", width = 8, height = 6, dpi = 300)

agtr1a_umap_zooom <- 
  FeaturePlot(combined_neuronal, features = "Agtr1a", raster = FALSE) + 
  coord_cartesian(xlim = c(-1,3), ylim = c(-1,2))
agtr1a_umap_zooom
ggsave("../../results/supplementary/agtr1a_umap_zoom.png", width = 5, height = 4, dpi = 300)

agtr1a_violin <- VlnPlot(combined_neuronal, features = "Agtr1a", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
agtr1a_violin
ggsave("../../results/supplementary/agtr1a_violin_plot.png", width = 12, height = 6, dpi = 300)

```


```{r make plots for Nxph4}

nxph4_umap_overall <- FeaturePlot(combined_neuronal, features = "Nxph4", raster = FALSE)
nxph4_umap_overall
ggsave("../../results/supplementary/nxph4_umap_full.png", width = 8, height = 6, dpi = 300)

nxph4_violin <- VlnPlot(combined_neuronal, features = "Nxph4", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
nxph4_violin
ggsave("../../results/supplementary/nxph4_violin_plot.png", width = 12, height = 6, dpi = 300)

# try again with map my cells labels
# class
Idents(combined_neuronal) <- combined_neuronal$ABC_class_name

nxph4_violin_class <- VlnPlot(combined_neuronal, features = "Nxph4", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
nxph4_violin_class
ggsave("../../results/supplementary/nxph4_violin_plot_class.png", width = 12, height = 6, dpi = 300)

# subclass
Idents(combined_neuronal) <- combined_neuronal$ABC_subclass_name

nxph4_violin_subclass <- VlnPlot(combined_neuronal, features = "Nxph4", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
nxph4_violin_subclass
ggsave("../../results/supplementary/nxph4_violin_plot_subclass.png", width = 12, height = 6, dpi = 300)

# hypomap
Idents(combined_neuronal) <- combined_neuronal$hypomap_predicted
  
nxph4_violin_hypomap <- VlnPlot(combined_neuronal, features = "Nxph4", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
nxph4_violin_hypomap
ggsave("../../results/supplementary/nxph4_violin_plot_hypomap.png", width = 12, height = 6, dpi = 300)

```

```{r also examine proportion of cells expressing nxph4 per assigned cell type}

Idents(combined_neuronal) <- combined_neuronal$cell_type

gene_expr <- GetAssayData(combined_neuronal, slot = "data")["Nxph4", ]

# Create a binary vector for expression
binary_expr <- ifelse(gene_expr > 0, 1, 0)

# Calculate percentage of expressing cells per group
expr_freq <- tapply(binary_expr, combined_neuronal$cell_type, mean) * 100

expr_freq <- as.data.frame(expr_freq)

expr_freq$cell_type <- rownames(expr_freq)
# make bar graph showing results
expr_freq <- expr_freq %>%
  arrange(desc(expr_freq)) %>%
  mutate(cell_type = factor(cell_type, levels = unique(cell_type)))

binary_nxph4_expr_prop <- ggplot(expr_freq, aes(x = cell_type, y = expr_freq)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))  +
  labs(x = "Cell Type", y = "Proportion of Nxph4\nExpressing Cells")

binary_nxph4_expr_prop

ggsave("../../results/supplementary/binary_nxph4_expr.png", width = 12, height = 6, dpi = 300)
```


```{r make plots for Myt1l}

myt1l_umap_overall <- FeaturePlot(combined_neuronal, features = "Myt1l", raster = FALSE)
myt1l_umap_overall
ggsave("../../results/supplementary/myt1l_umap_full.png", width = 8, height = 6, dpi = 300)

myt1l_violin <- VlnPlot(combined_neuronal, features = "Myt1l", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
myt1l_violin
ggsave("../../results/supplementary/myt1l_violin_plot.png", width = 12, height = 6, dpi = 300)
```






