---
title: "Map My Cells Probability Plot"
author: "Emma Jones"
date: "2025-07-31"
output: html_document
---

The purpose of this script is to generate the visualizations requested by some of our reviewers to clarify cell type assignments. First, we will make a plot of cell probability assignments from map my cells for each cluster, and then will make plots of some of our most notable genes in the paper.

We do not need to independently load in the map my cells results, as cell probability is included in the metadata of our main object.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up environment

```{r dependencies, warning = FALSE, message = FALSE}
# Change this to your package directory
.libPaths("/ref/jdlab/software/r-envs/simona/r-packages/4.2/")

# Load dependencies
library(dplyr)
library(Seurat)
library(patchwork)
library(Matrix)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(scCustomize)
library(patchwork)
library(qs)
library(viridis)
library(ggpubr)
library(SeuratDisk)
library(SeuratWrappers)
library(ComplexHeatmap)
library(stringr)
library(ggrastr)
library(parallel)
library(ggtree)
library(speckle)
library(limma)
library(gt)
library(data.table)
library(biomaRt)
library(tidyr)
library(compositions)
library(lisi)

set.seed(420)

source("../color_utils.R")
```

# Load in preprocessed data

```{r load rds}
combined_neuronal <- readRDS("../../data/250228_combined.neuronal_1-5_5pct.rds")
combined_neuronal
```
# Probability boxplots

First, we can use existing metadata to create probability boxplots.

```{r make bloxplots for probability}
# ggplot for class
class_probability_plot <- 
  ggplot(combined_neuronal@meta.data, aes(y=ABC_class_probability, x=ABC_class_name)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
      labs(x = "Class Names (33)", y = "Class Bootstrapping\nProbability")

class_probability_plot

ggsave("../../results/supplementary/class_probability_plot.png", width = 12, height = 4, dpi = 300)

# ggplot for subclass
subclass_probability_plot <- 
  ggplot(combined_neuronal@meta.data, aes(y=ABC_subclass_probability, x=ABC_subclass_name)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
      labs(x = "Subclass Names (274)", y = "Subclass Bootstrapping\nProbability")

subclass_probability_plot

ggsave("../../results/supplementary/subclass_probability_plot.png", width = 12, height = 4, dpi = 300)

```

# UMAP FeaturePlots of Agtr1a, Nxph4, and Myt1l

```{r make plots for Agtr1a}

agtr1a_umap_overall <- FeaturePlot(combined_neuronal, features = "Agtr1a", raster = FALSE)
agtr1a_umap_overall
ggsave("../../results/supplementary/agtr1a_umap_full.png", width = 8, height = 6, dpi = 300)

agtr1a_umap_zooom <- 
  FeaturePlot(combined_neuronal, features = "Agtr1a", raster = FALSE) + 
  coord_cartesian(xlim = c(-1,3), ylim = c(-1,2))
agtr1a_umap_zooom
ggsave("../../results/supplementary/agtr1a_umap_zoom.png", width = 5, height = 4, dpi = 300)

agtr1a_violin <- VlnPlot(combined_neuronal, features = "Agtr1a", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
agtr1a_violin
ggsave("../../results/supplementary/agtr1a_violin_plot.png", width = 12, height = 6, dpi = 300)

```


```{r make plots for Nxph4}

nxph4_umap_overall <- FeaturePlot(combined_neuronal, features = "Nxph4", raster = FALSE)
nxph4_umap_overall
ggsave("../../results/supplementary/nxph4_umap_full.png", width = 8, height = 6, dpi = 300)

nxph4_violin <- VlnPlot(combined_neuronal, features = "Nxph4", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
nxph4_violin
ggsave("../../results/supplementary/nxph4_violin_plot.png", width = 12, height = 6, dpi = 300)
```

```{r make plots for Myt1l}

myt1l_umap_overall <- FeaturePlot(combined_neuronal, features = "Myt1l", raster = FALSE)
myt1l_umap_overall
ggsave("../../results/supplementary/myt1l_umap_full.png", width = 8, height = 6, dpi = 300)

myt1l_violin <- VlnPlot(combined_neuronal, features = "Myt1l", raster = FALSE) + NoLegend() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) 
myt1l_violin
ggsave("../../results/supplementary/myt1l_violin_plot.png", width = 12, height = 6, dpi = 300)
```






