---
title: "Moxd1 OSCA Analyses"
author: "Emma Jones"
date: "2025-08-01"
output: html_document
---

The purpose of this script is to use a more conservative approach to testing cell abundance between two groups (sex and genotype). I am primarily following the statistical and code framework provided by the following tutorial https://bioconductor.org/books/3.13/OSCA.multisample/differential-abundance.html

There was also a request to validate the findings that moxd1 positive neurons are more prominent in males than females, so I will also perform an additional analysis with "moxd1-positive" and "moxd1-negative" counts for each sample.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Set up environment

```{r dependencies, warning = FALSE, message = FALSE}
# Change this to your package directory
.libPaths("/ref/jdlab/software/r-envs/simona/r-packages/4.2/")

# Load dependencies
library(dplyr)
library(Seurat)
library(patchwork)
library(Matrix)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(scCustomize)
library(patchwork)
library(qs)
library(viridis)
library(ggpubr)
library(SeuratDisk)
library(SeuratWrappers)
library(ComplexHeatmap)
library(stringr)
library(ggrastr)
library(parallel)
library(ggtree)
library(speckle)
library(limma)
library(gt)
library(data.table)
library(biomaRt)
library(tidyr)
library(compositions)
library(lisi)
library(edgeR)

set.seed(420)

source("../color_utils.R")
```

# Load in preprocessed data

```{r load rds}
combined_neuronal <- readRDS("../../data/250228_combined.neuronal_1-5_5pct.rds")
combined_neuronal
```


# Re-assess cell type abundances per animal

```{r get abundances table}

metadata <- combined_neuronal@meta.data

abundances <- table(metadata$cell_type, metadata$sample)
abundances <- unclass(abundances) 
head(abundances)

```


```{r get data formatted for edgeR}

sample_metadata <- metadata %>%
    group_by(sample) %>%
    reframe(
      sex = dplyr::first(sex),
      genotype = dplyr::first(genotype),
      learner = dplyr::first(Sac_Learner)
    )

y.ab <- DGEList(abundances, samples=sample_metadata)
y.ab

```

here we design our model, the options are to have ~sex + genotype, ~sex, ~genotype, or ~genotype + sex, ~learner, or ~sex + learner

```{r build model}

design <- model.matrix(~factor(learner), y.ab$samples)

y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)

plotBCV(y.ab, cex=1)

```

```{r examine QL dispersion}

fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

summary(fit.ab$df.prior)

# Plot the quasi-likelihood dispersion
plotQLDisp(fit.ab, cex=1)

```

```{r run statistical test}

res <- glmQLFTest(fit.ab, coef=ncol(design))

topTags(res, n=10)

summary(decideTests(res))

```
Nothing is significant for sex and genotype with OSCA here. Learner vs non-learner is significant when sex is not included in the model, but when sex is added the results are no longer significant after FDR correction.

## Try again with normalization factors

```{r normalizaton factors}
y.ab2 <- calcNormFactors(y.ab)
y.ab2$samples$norm.factors

```
```{r run test}
y.ab2 <- estimateDisp(y.ab2, design, trend="none")
fit.ab2 <- glmQLFit(y.ab2, design, robust=TRUE, abundance.trend=FALSE)
res2 <- glmQLFTest(fit.ab2, coef=ncol(design))
topTags(res2, n=10)

summary(decideTests(res2))

```

calcNormfactors does not work either.

But, let's try this method for Moxd1 neurons, as defined by Simona.

# Moxd1-specific analysis

## Simona Moxd1 code

To reiterate, this code is what Simona ran:

So any counts at all were considered expressing vs non-expressing.

```{r moxd1 violin in non-zero nucs}
gene_expr <- GetAssayData(combined_neuronal, slot = "data")["Moxd1", ]

# Create a binary vector for expression
binary_expr <- ifelse(gene_expr > 0, 1, 0)

# Calculate percentage of expressing cells per group
expr_freq <- tapply(binary_expr, combined_neuronal$sex, mean) * 100

# Plot non-zero values only
VlnPlot_scCustom(combined_neuronal[, gene_expr > 0],
  group.by = "sex",
  features = "Moxd1",
  pt.size = 1
) +
  ggtitle(paste0(
    "Moxd1 expression\n",
    "% expressing cells: F=", round(expr_freq["F"], 1),
    "%, M=", round(expr_freq["M"], 1), "%"
  ))
```
She than ran a wilcoxon test on the gene expression values for each cell, by sex. This is just above general significance thresholds, at 0.06744 and matches the results of n.s.

```{r wilcoxon test Moxd1 GEX}
test_data <- data.frame(
  Moxd1 = gene_expr,
  sex = combined_neuronal$sex
)

# Perform Wilcoxon test on non-zero values
wilcox_test <- wilcox.test(
  Moxd1 ~ sex,
  data = test_data[test_data$Moxd1 > 0, ],
  exact = FALSE
)

print(wilcox_test)
```
Now simona runs the proportions of expressing cells (any counts). This was found to be significant using her method.

```{r}
# For logit transformation

# Define logit function
logit <- function(p) log(p / (1 - p))

moxd1_data <- data.frame(
  sample = combined_neuronal$sample,
  sex = combined_neuronal$sex,
  Moxd1 = GetAssayData(combined_neuronal, slot = "data")["Moxd1", ]
)

# First calculate proportions by sample
moxd1_props <- moxd1_data %>%
  group_by(sample, sex) %>%
  summarize(
    total_cells = n(),
    expressing_cells = sum(Moxd1 > 0),
    proportion = expressing_cells / total_cells,
    .groups = "drop"
  )

# Logit transform the proportions for statistics
moxd1_props$transformed_prop <- logit(pmin(pmax(moxd1_props$proportion, 0.001), 0.999))
moxd1_props$standardized_prop <- scale(moxd1_props$transformed_prop)

# Fit linear model on transformed data
fit <- lm(standardized_prop ~ sex, data = moxd1_props)
summary_stats <- summary(fit)

# Calculate means and SE for untransformed proportions
sex_stats_raw <- moxd1_props %>%
  group_by(sex) %>%
  summarize(
    mean_prop = mean(proportion),
    se = sd(proportion) / sqrt(n()),
    .groups = "drop"
  )

# Plot with untransformed proportions
ggplot() +
  geom_bar(
    data = sex_stats_raw,
    aes(x = sex, y = mean_prop, fill = sex),
    stat = "identity",
    alpha = 0.8,
    width = 0.5
  ) +
  geom_errorbar(
    data = sex_stats_raw,
    aes(
      x = sex,
      ymin = mean_prop - se,
      ymax = mean_prop + se
    ),
    width = 0.2
  ) +
  geom_point(
    data = moxd1_props,
    aes(x = sex, y = proportion),
    position = position_jitter(width = 0.2),
    size = 1
  ) +
  scale_fill_manual(values = palette_sex) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.line = element_line(color = "black"),
    axis.text = element_text(color = "black", size = 12),
    axis.title = element_text(size = 14)
  ) +
  ylab("Proportion of Moxd1+ Nuclei") +
  xlab("Sex")

# Print model results (from transformed data)
print(summary_stats)
```

So, for validation using cell abundance, I will look at Moxd1 expression between males and females as binary presence or absence counts (cell abundance). This presence of absence is what what used to calculate the proportions in the plot above. I will not use expression values, as these were already used for Simona's DEG analysis and Wilcoxon rank sum test.


## New Moxd1 analysis

```{r get moxd1-positive abundances}
gene_expr <- GetAssayData(combined_neuronal, slot = "data")["Moxd1", ]

# Create a binary vector for expression
binary_expr <- ifelse(gene_expr > 0, 1, 0)

metadata$binary_expr <- binary_expr

# try other method of counts
moxd1_counts <- as.data.frame(table(metadata$sample, metadata$binary_expr))
colnames(moxd1_counts) <- c("sample", "binary_expr", "Freq")
# Reshape: make binary_expr values into columns: 0 = negative, 1 = positive
moxd1_counts <- reshape(moxd1_counts, idvar = "sample", timevar = "binary_expr", direction = "wide")
colnames(moxd1_counts) <- c("sample", "neg", "pos")

# Ensure samples line up with metadata
moxd1_counts <- merge(moxd1_counts, unique(metadata[, c("sample", "sex")]), by = "sample")

# Create counts matrix for edgeR
count_matrix <- t(as.matrix(moxd1_counts[, c("neg", "pos")]))
colnames(count_matrix) <- moxd1_counts$sample

count_matrix
```

```{r prepare data and design}
group <- factor(moxd1_counts$sex)
dge <- DGEList(counts = count_matrix, group = group)

dge

design <- model.matrix(~ group)

dge <- estimateDisp(dge, design)
fit <- glmQLFit(dge, design)

plotBCV(dge)
plotQLDisp(fit)

res <- glmQLFTest(fit, coef=2)  # coef=2 tests the 'group' variable
topTags(res)

summary(decideTests(res))
```
here, the result is just barely above an FDR cutoff of 0.1. maybe I can do a proportion test

```{r run prop test}
# First compute binary expression
metadata$Moxd1_positive <- ifelse(GetAssayData(combined_neuronal, slot = "data")["Moxd1", ] > 0, 1, 0)

# Count Moxd1+ and total cells per sample
table_by_sample <- aggregate(Moxd1_positive ~ sample, data = metadata, sum)
table_by_sample$total_cells <- as.numeric(table(metadata$sample))

# Merge with sex
sample_sex <- unique(metadata[, c("sample", "sex")])
table_by_sample <- merge(table_by_sample, sample_sex, by = "sample")

# Aggregate by sex
by_sex <- table_by_sample |>
  group_by(sex) |>
  summarise(
    moxd1_pos = sum(Moxd1_positive),
    total = sum(total_cells)
  )

# Run proportion test
prop.test(x = by_sex$moxd1_pos, n = by_sex$total)

```

Great, I was able to run a prop.test (asymptotic chi-square) and get p = 4.342e-08, so it is significant per-group.

can run a slightly different logistic regression

```{r run logistic regression}

table_by_sample$prop <- table_by_sample$Moxd1_positive / table_by_sample$total_cells

glm_model <- glm(
  cbind(Moxd1_positive, total_cells - Moxd1_positive) ~ sex,
  family = binomial,
  data = table_by_sample
)

summary(glm_model)

```

The logistic regression is also very significant, so we can feel more confident that "any" expression is higher in counts in male samples than females, but is this biologically meaningful? should there be a minimum expression cutoff and do males still have higher counts with a threshold in place?

```{r kernel density estimation}

ggplot(data = moxd1_props, aes(x = proportion, fill = sex, color = sex, alpha = 0.5)) +
  geom_density()

moxd1_props$genotype <- substr(moxd1_props$sample, 1, 3)
moxd1_props$genotype

ggplot(data = moxd1_props, aes(x = proportion, fill = sex, color = sex, alpha = 0.5)) +
  geom_density() +
  facet_wrap(~genotype)

```


```{r make histogram for moxd1 expression}
moxd1_data_no0 <- moxd1_data[moxd1_data$Moxd1 > 0,]

ggplot(data = moxd1_data_no0, aes(x = Moxd1, fill = sex, color = sex, alpha = 0.5)) +
  geom_histogram() +
  facet_wrap(~sample)

```

# Analyze cell count variability
This will include:
* Histograms of cell counts per labeled cell type
* "Plot a CV of the proportions vs. cell count plot and see if low abundant cells have too much variance in the estimate proportion to include them"
* plot CV of counts over mean counts

```{r examine cell abundances by histogram}

abundances_df <- as.data.frame(abundances)

for(cell_type in unique(metadata$cell_type)) {
  
  cell_type_safe <- gsub("/", "-", cell_type)
  
  png(paste("../../results/supplementary/" , cell_type_safe, "histogram.png"), width = 800, height = 600)
  
hist(as.numeric(abundances_df[cell_type,]), main = paste("Cell Counts of", cell_type))

dev.off()
  
}

for(cell_type in unique(metadata$cell_type)) {
  
hist(as.numeric(abundances_df[cell_type,]), main = paste("Cell Counts of", cell_type))
  
}

```

From these histograms, I think some of the rarest cell populations are too small to appropriately asses proportions or otherwise: Scsg Gabr2 Gaba, AD Serpinb7 Glut, LHA pmch glut, and SNc-VTA-RAmb Foxa1 Dopa

Next, I'll make a summary table of some stats.

```{r make a df of cv of counts, mean of counts, median of counts, and cv of proportion}
library(matrixStats)

abundance_summaries <- data.frame(matrix(nrow = nrow(abundances_df), ncol = 4))

rownames(abundance_summaries) <- rownames(abundances_df)

colnames(abundance_summaries) <- c("mean_counts", "median_counts", "cv_counts", "cv_prop")

abundance_summaries$mean_counts <- rowMeans(abundances_df)

abundance_summaries$median_counts <- rowMedians(as.matrix(abundances_df))

abundance_summaries$cv_counts <- rowSds(as.matrix(abundances_df)) / rowMeans(abundances_df)

abundance_summaries$sum_counts <- rowSums(abundances_df)

all_props <- abundances_df / colSums(abundances_df)

abundance_summaries$cv_prop <- rowSds(as.matrix(all_props)) / rowMeans(all_props)

abundance_summaries$min_counts <- rowMins(as.matrix(abundances_df))

abundance_summaries$max_counts <- rowMaxs(as.matrix(abundances_df))

```



```{r make lots of scatterplots!}
# prop cv and count cv with mean
ggplot(abundance_summaries, aes(x=cv_prop, y=mean_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_prop_mean.png")

ggplot(abundance_summaries, aes(x=cv_counts, y=mean_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_count_mean.png")

# prop cv and count cv with median
ggplot(abundance_summaries, aes(x=cv_prop, y=median_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_prop_median.png")

ggplot(abundance_summaries, aes(x=cv_counts, y=median_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_count_median.png")

# prop cv and count cv with sum
ggplot(abundance_summaries, aes(x=cv_prop, y=sum_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_prop_sum.png")

ggplot(abundance_summaries, aes(x=cv_counts, y=sum_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_count_sum.png")

# prop cv and count cv with min
ggplot(abundance_summaries, aes(x=cv_prop, y=min_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_prop_min.png")

ggplot(abundance_summaries, aes(x=cv_counts, y=min_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_count_min.png")

# prop cv and count cv with max
ggplot(abundance_summaries, aes(x=cv_prop, y=max_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_prop_max.png")

ggplot(abundance_summaries, aes(x=cv_counts, y=max_counts)) +
  geom_point()

ggsave("../../results/supplementary/cv_count_max.png")

```


It seems that median best represents the trend that as median counts decrease, CV increases. The best cutoff to use as minimum expression seems to be around 25, but this is still up for discussion and interpretation.










