---
title: "Analysis of IEG Activation in tuberal hypothalamus of MYT1L mice after social operant conditioning"
author: "Simona Sarafinovska"
date: "2025-02-13"
output:
---

# Set up environment
```{r message=FALSE, warning=TRUE}
# Change this to your directory
.libPaths('/ref/jdlab/software/r-envs/simona/r-packages/4.2/')

library(dplyr)
library(Seurat)
library(ggplot2)
library(tidyr)
library(patchwork)
library(corrplot)
library(tidyverse)
library(scCustomize)

set.seed(420)

source("color_utils.R")
```


# Load in data
```{r load in data}
combined_neuronal <- readRDS("../data/250228_combined.neuronal_1-5_5pct.rds")
combined_neuronal
```

```{r renaming cell_types for saving}
# Saving scatterplots won't work if you have "/" in cluster name
combined_neuronal@meta.data$cell_type <- as.factor(gsub("/", "-", combined_neuronal@meta.data$cell_type))
```

```{r subset for wt only}
# Subset the Seurat object to keep only WT cells
wt_only <- subset(combined_neuronal, subset = genotype == "WT")

# Verify the subsetting worked correctly
table(wt_only$genotype)
```

```{r drop unused sample levels}
# Drop unused factor levels from the sample column
wt_only$sample <- droplevels(wt_only$sample)

# Verify that only WT samples remain
levels(wt_only$sample)
```


# Fos/Jun/Arc expression plots
```{r fos expression violin plot}
gene_expr <- GetAssayData(wt_only, slot = "data")["Fos",]

# Create a binary vector for expression
binary_expr <- ifelse(gene_expr > 0, 1, 0)

# Calculate percentage of expressing cells per group
expr_freq <- tapply(binary_expr, wt_only$Sac_Learner, mean) * 100

VlnPlot_scCustom(wt_only[,gene_expr > 0],
                 group.by = "Sac_Learner",
                 features = "Fos",
                 colors_use = palette_learner,
                 pt.size = 1) +
  geom_boxplot(width=0.2, fill=NA, color="white", alpha=1, outlier.shape = NA) + 
  ggtitle(paste0("Fos expression\n",
                 "% expressing cells: Learner=", round(expr_freq["Learner"], 1), 
                 "%, Non_Learner=", round(expr_freq["Non_Learner"], 1), "%"))

ggsave(filename = '../results/IEGactivity_WTonly/250214_FosGEX_bylearner.pdf',
       device = pdf,
       width = 6,
       height = 8)
```

```{r wilcoxon test Fos GEX}
test_data <- data.frame(
  Fos = gene_expr,
  learner = wt_only$Sac_Learner
)

# Perform Wilcoxon test on non-zero values
wilcox_test <- wilcox.test(
  Fos ~ learner, 
  data = test_data[test_data$Fos > 0,], 
  exact = FALSE
)

print(wilcox_test)

# Calculate medians for direction
group_medians <- tapply(test_data[test_data$Fos > 0,]$Fos, 
                       test_data[test_data$Fos > 0,]$learner, 
                       median)

# Calculate effect size (r = Z/sqrt(N))
# First get Z score from wilcox test
z_score <- qnorm(wilcox_test$p.value/2) # divided by 2 for two-tailed test
n <- nrow(test_data[test_data$Fos > 0,])
effect_size_r <- abs(z_score/sqrt(n))

# Print results
print("Medians:")
print(group_medians)
print(paste("Effect size (r):", round(effect_size_r, 3)))
```

```{r jun expression violin plot}
gene_expr <- GetAssayData(wt_only, slot = "data")["Jun",]

# Create a binary vector for expression
binary_expr <- ifelse(gene_expr > 0, 1, 0)

# Calculate percentage of expressing cells per group
expr_freq <- tapply(binary_expr, wt_only$Sac_Learner, mean) * 100

VlnPlot_scCustom(wt_only[,gene_expr > 0],
                 group.by = "Sac_Learner",
                 features = "Jun",
                 colors_use = palette_learner,
                 pt.size = 1) +
  geom_boxplot(width=0.2, fill=NA, color="white", alpha=1, outlier.shape = NA) + 
  ggtitle(paste0("Jun expression\n",
                 "% expressing cells: Learner=", round(expr_freq["Learner"], 1), 
                 "%, Non_Learner=", round(expr_freq["Non_Learner"], 1), "%"))

ggsave(filename = '../results/IEGactivity_WTonly/250214_JunGEX_bylearner.pdf',
       device = pdf,
       width = 6,
       height = 8)
```

```{r wilcoxon test Jun GEX}
test_data <- data.frame(
  Jun = gene_expr,
  learner = wt_only$Sac_Learner
)

# Perform Wilcoxon test on non-zero values
wilcox_test <- wilcox.test(
  Jun ~ learner, 
  data = test_data[test_data$Jun > 0,], 
  exact = FALSE
)

print(wilcox_test)

# Calculate medians for direction
group_medians <- tapply(test_data[test_data$Jun > 0,]$Jun, 
                       test_data[test_data$Jun > 0,]$learner, 
                       median)

# Calculate effect size (r = Z/sqrt(N))
# First get Z score from wilcox test
z_score <- qnorm(wilcox_test$p.value/2) # divided by 2 for two-tailed test
n <- nrow(test_data[test_data$Jun > 0,])
effect_size_r <- abs(z_score/sqrt(n))

# Print results
print("Medians:")
print(group_medians)
print(paste("Effect size (r):", round(effect_size_r, 3)))
```

```{r arc expression violin plot}
gene_expr <- GetAssayData(wt_only, slot = "data")["Arc",]

# Create a binary vector for expression
binary_expr <- ifelse(gene_expr > 0, 1, 0)

# Calculate percentage of expressing cells per group
expr_freq <- tapply(binary_expr, wt_only$Sac_Learner, mean) * 100

VlnPlot_scCustom(wt_only[,gene_expr > 0],
                 group.by = "Sac_Learner",
                 features = "Arc",
                 colors_use = palette_learner,
                 pt.size = 1) +
  geom_boxplot(width=0.2, fill=NA, color="white", alpha=1, outlier.shape = NA) + 
  ggtitle(paste0("Arc expression\n",
                 "% expressing cells: Learner=", round(expr_freq["Learner"], 1), 
                 "%, Non_Learner=", round(expr_freq["Non_Learner"], 1), "%"))

ggsave(filename = '../results/IEGactivity_WTonly/250214_ArcGEX_bylearner.pdf',
       device = pdf,
       width = 6,
       height = 8)
```

```{r wilcoxon test Arc GEX}
test_data <- data.frame(
  Arc = gene_expr,
  learner = wt_only$Sac_Learner
)

# Perform Wilcoxon test on non-zero values
wilcox_test <- wilcox.test(
  Arc ~ learner, 
  data = test_data[test_data$Arc > 0,], 
  exact = FALSE
)

print(wilcox_test)

# Calculate medians for direction
group_medians <- tapply(test_data[test_data$Arc > 0,]$Arc, 
                       test_data[test_data$Arc > 0,]$learner, 
                       median)

# Calculate effect size (r = Z/sqrt(N))
# First get Z score from wilcox test
z_score <- qnorm(wilcox_test$p.value/2) # divided by 2 for two-tailed test
n <- nrow(test_data[test_data$Arc > 0,])
effect_size_r <- abs(z_score/sqrt(n))

# Print results
print("Medians:")
print(group_medians)
print(paste("Effect size (r):", round(effect_size_r, 3)))
```

# Stats on Pct activated by Sample

```{r filtered iegs list}
filtered_iegs <- c("Fos", "Fosb", "Fosl2", "Jun", "Junb", "Jund", "Egr1", "Egr3", "Egr4", "Nr4a1", "Nr4a2", "Nr4a3", "Arc", "Rgs2", "Plk2", "Bdnf", "Inhba", "Nptx2", "Plat", "Nrn1", "Dusp1", "Dusp5", "Dusp6", "Pcdh8", "Trib1", "Gem", "Btg2", "Npas4", "Rasd1", "Gadd45g", "Ppp1r15a", "Fbxo33", "Srf", "Mcl1", "Bhlhe40", "Nfkbia", "Klf6", "Klf10", "Nfkbiz", "Tnfaip3", "Hbegf", "Ldlr", "F3", "Csrnp1", "Zfp36l2", "Irf1", "Cd83", "Map3k8", "Pim1", "Noct", "Ifit2", "Vcam1", "Ackr4", "Marcksl1", "Nfkbid", "Ikbke", "Zfp36l1", "Hes1", "Wee1", "Sik1", "Ier5", "Id2")
```

## by Sex

```{r fishers pct activated by sex}
Idents(wt_only) <- wt_only@meta.data$cell_type

# Fetch expression data for all IEGs
ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Sex_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_df %>%
  group_by(Cluster, Sex, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Perform Fisher's exact test on activated percentages between Sexs within each cluster
cluster_ieg_Sex_comparison <- cluster_ieg_Sex_sample_percentages %>%
  group_by(Cluster) %>%
  summarise(
    Activated_Percentage_M_mean = mean(Activated_Percentage[Sex == "M"]),
    Activated_Percentage_F_mean = mean(Activated_Percentage[Sex == "F"]),
    Activated_Percentage_Difference = Activated_Percentage_M_mean - Activated_Percentage_F_mean,
    p_value = fisher.test(
      x = matrix(
        c(
          sum(Activated_Cells[Sex == "M"]),
          sum(Cluster_Cells[Sex == "M"] - Activated_Cells[Sex == "M"]),
          sum(Activated_Cells[Sex == "F"]),
          sum(Cluster_Cells[Sex == "F"] - Activated_Cells[Sex == "F"])
        ),
        nrow = 2
      )
    )$p.value
  )

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values <- p.adjust(cluster_ieg_Sex_comparison$p_value, method = "BH")

# Add adjusted p-values to the results data frame
cluster_ieg_Sex_comparison$AdjustedPValue <- adjusted_p_values

cluster_ieg_Sex_comparison <- cluster_ieg_Sex_comparison %>%
    arrange(AdjustedPValue)

# Print the updated results
print(cluster_ieg_Sex_comparison)

# # Write the results to a CSV file
write.csv(cluster_ieg_Sex_comparison, file = "../results/IEGactivity_WTonly/250303_activationpercent_bysex.csv", row.names = FALSE)
```

## by Learner 

```{r fishers pct activated by learner}
Idents(wt_only) <- wt_only@meta.data$cell_type

# Fetch expression data for all IEGs
#ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Learner_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Learner = wt_only$Sac_Learner,
  Sample = wt_only$sample,
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_Learner_sample_percentages <- cluster_ieg_Learner_sample_df %>%
  group_by(Cluster, Learner, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Perform Fisher's exact test on activated percentages between Sexs within each cluster
cluster_ieg_Learner_comparison <- cluster_ieg_Learner_sample_percentages %>%
  group_by(Cluster) %>%
  summarise(
    Activated_Percentage_Learner_mean = mean(Activated_Percentage[Learner == "Learner"]),
    Activated_Percentage_Non_Learner_mean = mean(Activated_Percentage[Learner == "Non_Learner"]),
    Activated_Percentage_Difference = Activated_Percentage_Learner_mean - Activated_Percentage_Non_Learner_mean,
    p_value = fisher.test(
      x = matrix(
        c(
          sum(Activated_Cells[Learner == "Learner"]),
          sum(Cluster_Cells[Learner == "Learner"] - Activated_Cells[Learner == "Learner"]),
          sum(Activated_Cells[Learner == "Non_Learner"]),
          sum(Cluster_Cells[Learner == "Non_Learner"] - Activated_Cells[Learner == "Non_Learner"])
        ),
        nrow = 2
      )
    )$p.value
  )

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values <- p.adjust(cluster_ieg_Learner_comparison$p_value, method = "BH")

# Add adjusted p-values to the results data frame
cluster_ieg_Learner_comparison$AdjustedPValue <- adjusted_p_values

cluster_ieg_Learner_comparison <- cluster_ieg_Learner_comparison %>%
    arrange(AdjustedPValue)

# Print the updated results
print(cluster_ieg_Learner_comparison)

# Write the results to a CSV file
write.csv(cluster_ieg_Learner_comparison, file = "../results/IEGactivity_WTonly/250303_activationpercent_bylearner.csv", row.names = FALSE)
```

# Percentage bar plots

## by Sex
```{r pct activated by sex df}
# Fetch expression data for all IEGs
#ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, sex, and IEG expression
cluster_ieg_sex_df <- data.frame(
  Cluster = Idents(wt_only),
  Sample = wt_only@meta.data$sample,
  Sex = wt_only@meta.data$sex,
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster and sample
cluster_ieg_sex_percentages <- cluster_ieg_sex_df %>%
  group_by(Cluster, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100)  %>%
  ungroup()

# Convert Cluster to a factor with desired order of levels
cluster_ieg_sex_percentages$Cluster <- factor(cluster_ieg_sex_percentages$Cluster,
                                              levels = sort(as.character(levels(wt_only))))

cluster_ieg_sex_percentages$Sex <- gsub('.*_(\\w)_.*', '\\1', cluster_ieg_sex_percentages$Sample)
```

```{r pct activated by sex calculations}
#Determine mean average per cluster for sex
cluster_ieg_sex_percentages_norm <- cluster_ieg_sex_percentages %>%
  group_by(Cluster, Sex) %>%
  dplyr::mutate(Mean_Activated_Percentage = mean(Activated_Percentage)) %>%
  dplyr::mutate(SE_Mean = sd(Activated_Percentage)/sqrt(36)) 

WT_M_1_values <- cluster_ieg_sex_percentages_norm[cluster_ieg_sex_percentages_norm$Sample == "WT_M_1", c("Cluster", "Mean_Activated_Percentage")]
cluster_ieg_sex_percentages_norm <- merge(cluster_ieg_sex_percentages_norm, WT_M_1_values, by = "Cluster", all.x = TRUE)
colnames(cluster_ieg_sex_percentages_norm)[colnames(cluster_ieg_sex_percentages_norm) == "Mean_Activated_Percentage.y"] <- "Mean_Activated_Percentage_M"

cluster_ieg_sex_percentages_norm <- cluster_ieg_sex_percentages_norm %>%
  dplyr::mutate(Activated_Percentage_normalizedtoM = ((Activated_Percentage / Mean_Activated_Percentage_M) - 1)) 

# Make sure the WT mean is around 0
cluster_ieg_sex_percentages_norm %>%
  filter(Sex == "M") %>%
  group_by(Cluster) %>%
  summarize(mean = mean(Activated_Percentage_normalizedtoM))

cluster_ieg_sex_percentages_norm <- cluster_ieg_sex_percentages_norm %>%
  group_by(Cluster, Sex) %>%
  dplyr::mutate(Mean_Activated_Percentage_normalizedtoM = mean(Activated_Percentage_normalizedtoM)) %>%
  dplyr::mutate(SE_Mean_Activated_Percentage_normalizedtoM = sd(Activated_Percentage_normalizedtoM)/sqrt(36))
```

```{r pct activated by geno plot}
p1 <- ggplot(cluster_ieg_sex_percentages_norm, aes(x = Mean_Activated_Percentage.x, y = Cluster, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge",
           width = 0.7) +
  geom_errorbar(aes(y = Cluster, xmin = Mean_Activated_Percentage.x - SE_Mean, xmax = Mean_Activated_Percentage.x + SE_Mean),
                position = position_dodge(0.7),
                width = 0.2, size = 0.2) +
    geom_point(aes(x = Activated_Percentage, y = Cluster),
             position = position_dodge(width = 0.7),
             color = "black",
             size = 0.01) +
  labs(x = "Proportion", y = "Cluster") +
  theme_classic() +
  scale_fill_manual(values = palette_sex) +
  theme(legend.position = "right")

p2 <- ggplot(cluster_ieg_sex_percentages_norm, aes(x = Mean_Activated_Percentage_normalizedtoM, y = Cluster, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge",
           width = 0.7) +
  geom_errorbar(aes(y = Cluster, xmin = Mean_Activated_Percentage_normalizedtoM - SE_Mean_Activated_Percentage_normalizedtoM, xmax = Mean_Activated_Percentage_normalizedtoM + SE_Mean_Activated_Percentage_normalizedtoM),
                position = position_dodge(0.7),
                width = 0.2, size = 0.2) +
  labs(x = "Proportion", y = "") +
  theme_classic() +
  scale_fill_manual(values = palette_sex) +
  theme(axis.text.y = element_blank(),
        legend.position = "none")

wrap_plots(p1, p2)
ggsave('../results/IEGactivity_WTonly/250303-activationbySex_neuronal_1-5_5pct_50_1.8.pdf',
       device = pdf,
       width = 12,
       height = 8)
```

## by Learner
```{r pct activated by learner df}
# Fetch expression data for all IEGs
#ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, sex, and IEG expression
cluster_ieg_Learner_df <- data.frame(
  Cluster = Idents(wt_only),
  Sample = wt_only@meta.data$sample,
  Learner = wt_only@meta.data$Sac_Learner,
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster and sample
cluster_ieg_Learner_percentages <- cluster_ieg_Learner_df %>%
  group_by(Cluster, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    Learner = dplyr::first(Learner) 
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100)  %>%
  ungroup()
```

```{r pct activated by learner calculations}
#Determine mean average per cluster for sex
cluster_ieg_Learner_percentages_norm <- cluster_ieg_Learner_percentages %>%
  group_by(Cluster, Learner) %>%
  dplyr::mutate(Mean_Activated_Percentage = mean(Activated_Percentage)) %>%
  dplyr::mutate(SE_Mean = sd(Activated_Percentage)/sqrt(36)) 

WT_M_10_values <- cluster_ieg_Learner_percentages_norm[cluster_ieg_Learner_percentages_norm$Sample == "WT_M_10", c("Cluster", "Mean_Activated_Percentage")]
cluster_ieg_Learner_percentages_norm <- merge(cluster_ieg_Learner_percentages_norm, WT_M_10_values, by = "Cluster", all.x = TRUE)
colnames(cluster_ieg_Learner_percentages_norm)[colnames(cluster_ieg_Learner_percentages_norm) == "Mean_Activated_Percentage.y"] <- "Mean_Activated_Percentage_NL"

cluster_ieg_Learner_percentages_norm <- cluster_ieg_Learner_percentages_norm %>%
  dplyr::mutate(Activated_Percentage_normalizedtoNL= ((Activated_Percentage / Mean_Activated_Percentage_NL) - 1)) 

# Make sure the WT mean is around 0
cluster_ieg_Learner_percentages_norm %>%
  filter(Learner == "Non_Learner") %>%
  group_by(Cluster) %>%
  summarize(mean = mean(Activated_Percentage_normalizedtoNL))

cluster_ieg_Learner_percentages_norm <- cluster_ieg_Learner_percentages_norm %>%
  group_by(Cluster, Learner) %>%
  dplyr::mutate(Mean_Activated_Percentage_normalizedtoNL = mean(Activated_Percentage_normalizedtoNL)) %>%
  dplyr::mutate(SE_Mean_Activated_Percentage_normalizedtoNL = sd(Activated_Percentage_normalizedtoNL)/sqrt(36))
```

```{r pct activated by learner plot}
p1 <- ggplot(cluster_ieg_Learner_percentages_norm, aes(x = Mean_Activated_Percentage.x, y = Cluster, fill = Learner)) +
  geom_bar(stat = "identity", position = "dodge",
           width = 0.7) +
  geom_errorbar(aes(y = Cluster, xmin = Mean_Activated_Percentage.x - SE_Mean, xmax = Mean_Activated_Percentage.x + SE_Mean),
                position = position_dodge(0.7),
                width = 0.2, size = 0.2) +
  geom_point(aes(x = Activated_Percentage, y = Cluster),
             position = position_dodge(width = 0.7),
             color = "black",
             size = 0.01) +
  labs(x = "Proportion", y = "Cluster") +
  theme_classic() +
  scale_fill_manual(values = palette_learner) +
  theme(legend.position = "right")

p2 <- ggplot(cluster_ieg_Learner_percentages_norm, aes(x = Mean_Activated_Percentage_normalizedtoNL, y = Cluster, fill = Learner)) +
  geom_bar(stat = "identity", position = "dodge",
           width = 0.7) +
  geom_errorbar(aes(y = Cluster, xmin = Mean_Activated_Percentage_normalizedtoNL - SE_Mean_Activated_Percentage_normalizedtoNL, xmax = Mean_Activated_Percentage_normalizedtoNL + SE_Mean_Activated_Percentage_normalizedtoNL),
                position = position_dodge(0.7),
                width = 0.2, size = 0.2) +
  labs(x = "Proportion", y = "") +
  theme_classic() +
  scale_fill_manual(values = palette_learner) +
  theme(axis.text.y = element_blank(),
        legend.position = "none")

wrap_plots(p1, p2)
ggsave('../results/IEGactivity_WTonly/250303-activationbyLearner_neuronal_1-5_5pct_50_1.8.pdf',
       device = pdf,
       width = 12,
       height = 8)
```

```{r sig clusters only plotr}
significant_clusters <- cluster_ieg_Learner_comparison %>%
  filter(p_value < 0.1) %>%
  pull(Cluster) %>%
  as.character()

sig_cluster_ieg_Learner_percentages_norm <- cluster_ieg_Learner_percentages_norm %>%
  dplyr::filter(Cluster %in% significant_clusters) %>%
  arrange(match(Cluster, significant_clusters))

sig_p1 <- ggplot(sig_cluster_ieg_Learner_percentages_norm, aes(x = Mean_Activated_Percentage.x, y = Cluster, fill = Learner)) +
  geom_bar(stat = "identity", position = "dodge",
           width = 0.7) +
  geom_errorbar(aes(y = Cluster, xmin = Mean_Activated_Percentage.x - SE_Mean, xmax = Mean_Activated_Percentage.x + SE_Mean),
                position = position_dodge(0.7),
                width = 0.2, size = 0.2) +
  labs(x = "Proportion", y = "Cluster") +
  theme_classic() +
  scale_fill_manual(values =palette_learner) +
  geom_point(aes(x = Activated_Percentage, y = Cluster),
             position = position_dodge(width = 0.7),
             color = "black",
             size = 0.01) +
  theme(legend.position = "right")

sig_p1

ggsave('../results/IEGactivity_WTonly/250303-activationbyLearner_neuronal_1-5_5pct_50_1.8_sigonly.pdf',
       device = pdf,
       width = 5,
       height = 8)
```

# Regressions of Behavior by Pct activated
## Rewards
```{r regression rewards by pct activated}
Idents(wt_only) <- wt_only$cell_type

# Fetch expression data for all IEGs
ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  TotalRewards = wt_only$Avg_Total_Rewards, 
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_sample_percentages <- cluster_ieg_sample_df %>%
  group_by(Cluster, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    TotalRewards = dplyr::first(TotalRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

cluster_regression_results <- cluster_ieg_sample_percentages %>%
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(TotalRewards ~ Activated_Percentage, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    p_value = map_dbl(summary, ~ .x$coefficients[2, 4]),
    r_squared = map_dbl(summary, ~ .x$r.squared)
  ) %>%
  dplyr::mutate(
    Slope = map_dbl(coefficients, ~ .x[2]),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope, Intercept, p_value, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values <- p.adjust(cluster_regression_results$p_value, method = "BH")

# Add adjusted p-values to the results data frame
cluster_regression_results$AdjustedPValue <- adjusted_p_values
cluster_regression_results <- cluster_regression_results %>%
    arrange(AdjustedPValue)

# Print the updated results
print(cluster_regression_results)

# # Write the results to a CSV file
write.csv(cluster_regression_results, file = "../results/IEGactivity_WTonly/250303_rewards_byactivationpercent.csv", row.names = FALSE)
```

```{r scatters rewards by pct activated}
plot_cluster <- function(cluster, cluster_regression_results, cluster_ieg_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- cluster_regression_results[cluster_regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$TotalRewards, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Total Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l")
  
  # Fit line
  fit <- lm(TotalRewards ~ Activated_Percentage, data = cluster_df)
  
  # Add the fit line
  abline(fit, col = "black", lwd = 2)
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("R² =", round(cluster_results$r_squared, 3)),
    paste("p-value =", format.pval(cluster_results$p_value, digits = 3)),
    paste("Adj. p-value =", format.pval(cluster_results$AdjustedPValue, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = "black",
         cex = 1.2,
         bty = "n")
  
  # Add title with cluster name and coefficients
  title_text <- paste("Cluster", cluster, "\n",
                      "Intercept =", round(cluster_results$Intercept, 3), "\n",
                      "Slope =", round(cluster_results$Slope, 3))
  title(main = title_text, cex.main = 1.5)
}

# Get all unique cluster numbers
clusters <- unique(cluster_regression_results$Cluster)

# Create a directory to save the PDFs if it doesn't exist
dir.create("../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303", showWarnings = FALSE)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303/cluster_", cluster, "_plot.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster(cluster, cluster_regression_results, cluster_ieg_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'RewardsbyActivatedPercentage' directory.\n")
```

### with Sex
```{r regression rewards by pct activated and sex }
# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Sex_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  TotalRewards = wt_only$Avg_Total_Rewards, 
  ieg_expression
)
# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_df %>%
  group_by(Cluster, Sex, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    TotalRewards = dplyr::first(TotalRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()
# Perform linear regression for each cluster, skipping those with only one Sex level
regression_results <- cluster_ieg_Sex_sample_percentages %>%
  # First identify clusters with only one Sex level
  group_by(Cluster) %>%
  mutate(sex_levels_count = n_distinct(Sex)) %>%
  ungroup() %>%
  # Filter to keep only clusters with at least 2 Sex levels
  filter(sex_levels_count >= 2) %>%
  # Now proceed with your analysis on the filtered data
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(TotalRewards ~ Activated_Percentage + Sex, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    r_squared = map_dbl(summary, ~ .x$r.squared),
    p_value_activated_percentage = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 2 && ncol(coef_matrix) >= 4) .x$coefficients[2, 4] else NA_real_
    }),
    p_value_sex = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 3 && ncol(coef_matrix) >= 4) .x$coefficients[3, 4] else NA_real_
    })
  ) %>%
  dplyr::mutate(
    Slope_Activated_Percentage = map_dbl(coefficients, ~ {
      if (length(.x) >= 2) .x[2] else NA_real_
    }),
    Slope_Sex = map_dbl(coefficients, ~ {
      if (length(.x) >= 3) .x[3] else NA_real_
    }),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope_Activated_Percentage, Slope_Sex, Intercept,
         p_value_activated_percentage, p_value_sex, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values_activated <- p.adjust(regression_results$p_value_activated_percentage, method = "BH")
adjusted_p_values_sex <- p.adjust(regression_results$p_value_sex, method = "BH")

# Add adjusted p-values to the results data frame
regression_results$adjusted_p_values_activated <-adjusted_p_values_activated
regression_results$adjusted_p_values_sex <- adjusted_p_values_sex

regression_results <- regression_results %>%
    arrange(adjusted_p_values_activated)

# Print the updated results
print(regression_results)

# Write the results to a CSV file
write.csv(regression_results, file = "../results/IEGactivity_WTonly/250303_rewards_byactivationpercent_bysex.csv", row.names = FALSE)
```

```{r scatters rewards by pct activated with sex}
plot_cluster_withsex <- function(cluster, regression_results, cluster_ieg_Sex_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- regression_results[regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_Sex_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$TotalRewards, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l",
       col = ifelse(cluster_df$Sex == "M", "#4478AB", "#ED6677"))
  
  # Fit lines for all data, male, and female
  fit_all <- lm(TotalRewards ~ Activated_Percentage + Sex, data = cluster_df)
  fit_male <- lm(TotalRewards ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "M", ])
  fit_female <- lm(TotalRewards ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "F", ])
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_male, col = "#4478AB", lwd = 2)
  abline(fit_female, col = "#ED6677", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_male <- summary(fit_male)$r.squared
  r2_female <- summary(fit_female)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Male (R² =", round(r2_male, 3), ")"),
    paste("Female (R² =", round(r2_female, 3), ")"),
    paste("Activated % Slope =", round(cluster_results$Slope_Activated_Percentage, 3)),
    paste("Sex Slope =", round(cluster_results$Slope_Sex, 3)),
    paste("Intercept =", round(cluster_results$Intercept, 3)),
    paste("Activated % p-value =", format.pval(cluster_results$p_value_activated_percentage, digits = 3)),
    paste("Sex p-value =", format.pval(cluster_results$p_value_sex, digits = 3)),
    paste("Activated % Adj. p-value =", format.pval(cluster_results$adjusted_p_values_activated, digits = 3)),
    paste("Sex Adj. p-value =", format.pval(cluster_results$adjusted_p_values_sex, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#4478AB", "#ED6677", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")

} 

# Get all unique cluster numbers
clusters <- unique(regression_results$Cluster)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303/cluster_", cluster, "_plot_bySex.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster_withsex(cluster, regression_results, cluster_ieg_Sex_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'RewardsbyActivatedPercentage' directory.\n")
```

### with Learner 
```{r regression rewards by pct activated and learner}
# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Learner_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sample = wt_only$sample,
  Sac_Learner = wt_only$Sac_Learner, 
  TotalRewards = wt_only$Avg_Total_Rewards, 
  ieg_expression
)
# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sac_Learner, and sample
cluster_ieg_Learner_sample_percentages <- cluster_ieg_Learner_sample_df %>%
  group_by(Cluster, Sac_Learner, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    TotalRewards = dplyr::first(TotalRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

regression_results <- cluster_ieg_Learner_sample_percentages %>%
  # First identify clusters with only one Sac_Learner level
  group_by(Cluster) %>%
  mutate(learner_levels_count = n_distinct(Sac_Learner)) %>%
  ungroup() %>%
  # Filter to keep only clusters with at least 2 Sac_Learner levels
  filter(learner_levels_count >= 2) %>%
  # Now proceed with your analysis on the filtered data
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(TotalRewards ~ Activated_Percentage + Sac_Learner, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    r_squared = map_dbl(summary, ~ .x$r.squared),
    p_value_activated_percentage = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 2 && ncol(coef_matrix) >= 4) .x$coefficients[2, 4] else NA_real_
    }),
    p_value_sac_learner = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 3 && ncol(coef_matrix) >= 4) .x$coefficients[3, 4] else NA_real_
    })
  ) %>%
  dplyr::mutate(
    Slope_Activated_Percentage = map_dbl(coefficients, ~ {
      if (length(.x) >= 2) .x[2] else NA_real_
    }),
    Slope_Sac_Learner = map_dbl(coefficients, ~ {
      if (length(.x) >= 3) .x[3] else NA_real_
    }),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope_Activated_Percentage, Slope_Sac_Learner, Intercept,
         p_value_activated_percentage, p_value_sac_learner, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values_activated <- p.adjust(regression_results$p_value_activated_percentage, method = "BH")
adjusted_p_values_sac_learner <- p.adjust(regression_results$p_value_sac_learner, method = "BH")

# Add adjusted p-values to the results data frame
regression_results$adjusted_p_values_activated <-adjusted_p_values_activated
regression_results$adjusted_p_values_sac_learner <- adjusted_p_values_sac_learner

regression_results <- regression_results %>%
    arrange(adjusted_p_values_activated)

# Print the updated results
print(regression_results)

# Write the results to a CSV file
write.csv(regression_results, file = "../results/IEGactivity_WTonly/250303_rewards_byactivationpercent_bysaclearner.csv", row.names = FALSE)
```

```{r scatters rewards by pct activated with learner}
plot_cluster_withSac_Learner <- function(cluster, regression_results, cluster_ieg_Learner_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- regression_results[regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_Learner_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$TotalRewards, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l",
       col = ifelse(cluster_df$Sac_Learner == "Learner", "#547B80", "#D1D3D4"))
  
  # Fit lines for all data, male, and female
  fit_all <- lm(TotalRewards ~ Activated_Percentage + Sac_Learner, data = cluster_df)
  fit_learner <- lm(TotalRewards ~ Activated_Percentage, data = cluster_df[cluster_df$Sac_Learner == "Learner", ])
  fit_nonlearner <- lm(TotalRewards ~ Activated_Percentage, data = cluster_df[cluster_df$Sac_Learner == "Non_Learner", ])
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_learner, col = "#547B80", lwd = 2)
  abline(fit_nonlearner, col = "#D1D3D4", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_learner <- summary(fit_learner)$r.squared
  r2_nonleaner <- summary(fit_nonlearner)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Learner (R² =", round(r2_learner, 3), ")"),
    paste("Non-Learner (R² =", round(r2_nonleaner, 3), ")"),
    paste("Activated % Slope =", round(cluster_results$Slope_Activated_Percentage, 3)),
     paste("Sac_Learner Slope =", round(cluster_results$Slope_Sac_Learner, 3)), 
    paste("Intercept =", round(cluster_results$Intercept, 3)),
    paste("Activated % p-value =", format.pval(cluster_results$p_value_activated_percentage, digits = 3)),
    paste("Sac_Learner p-value =", format.pval(cluster_results$p_value_sac_sLearner, digits = 3)),
    paste("Activated % Adj. p-value =", format.pval(cluster_results$adjusted_p_values_activated, digits = 3)),
    paste("Sac_Learner Adj. p-value =", format.pval(cluster_results$adjusted_p_values_sac_learner, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#547B80", "#D1D3D4", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")

} 

# Get all unique cluster numbers
clusters <- unique(regression_results$Cluster)
problem_clusters <- c("Hist Gaba", "PO-MEA Slc17a7 Glut")
clusters_to_plot <- clusters[!clusters %in% problem_clusters] # ablines wouldn't fit

# Iterate through clusters
for (cluster in clusters_to_plot) {
  print(cluster)
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303/cluster_", cluster, "_plot_bySac_Learner.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster_withSac_Learner(cluster, regression_results, cluster_ieg_Learner_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'RewardsbyActivatedPercentage' directory.\n")
```

```{r}
plot_cluster_withSac_Learner_no_abline <- function(cluster, regression_results, cluster_ieg_Learner_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- regression_results[regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_Learner_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot - just the scatter points, no regression lines
  plot(cluster_df$Activated_Percentage, cluster_df$TotalRewards, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage, na.rm = TRUE) + 1),         
       bty = "l",
       col = ifelse(!is.na(cluster_df$Sac_Learner) & cluster_df$Sac_Learner == "Learner", "#547B80", "#D1D3D4"))
  
  # Create legend with just the regression statistics, no R² values
  if(nrow(cluster_results) > 0) {
    legend_labels <- c(
      "Learner",
      "Non-Learner",
      paste("Activated % Slope =", round(cluster_results$Slope_Activated_Percentage, 3)),
      paste("Sac_Learner Slope =", round(cluster_results$Slope_Sac_Learner, 3)),
      paste("Intercept =", round(cluster_results$Intercept, 3)),
      paste("Activated % p-value =", format.pval(cluster_results$p_value_activated_percentage, digits = 3)),
      paste("Sac_Learner p-value =", format.pval(cluster_results$p_value_sac_learner, digits = 3)),
      paste("Activated % Adj. p-value =", format.pval(cluster_results$adjusted_p_values_activated, digits = 3)),
      paste("Sac_Learner Adj. p-value =", format.pval(cluster_results$adjusted_p_values_sac_learner, digits = 3))
    )
    
    legend("topleft", 
           legend = legend_labels,
           col = c("#547B80", "#D1D3D4", rep("black", 7)),
           pch = c(16, 16, rep(NA, 7)),
           cex = 1.2,
           bty = "n")
  }
}

# Iterate through clusters
for (cluster in problem_clusters) {
  print(cluster)
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303/cluster_", cluster, "_plot_bySac_Learner_noabline.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster_withSac_Learner_no_abline(cluster, regression_results, cluster_ieg_Learner_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'RewardsbyActivatedPercentage' directory.\n")
```

### Control: by Pct Activity in All Clusters
```{r scatter rewards by pct activity in all neurons}
# First calculate the activation percentage across all neurons for each sample
overall_sample_percentages <- cluster_ieg_sample_df %>%
  group_by(Sample) %>%
  summarise(
    Total_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    TotalRewards = dplyr::first(TotalRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Total_Cells * 100) %>%
  ungroup()

# Create a plotting function for the control
plot_control <- function(sample_data) {
  # Fit regression line
  fit <- lm(TotalRewards ~ Activated_Percentage, data = sample_data)
  fit_summary <- summary(fit)
  
  # Create the plot
  plot(sample_data$Activated_Percentage, sample_data$TotalRewards, 
       main = "All Neurons",
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       #xlim = c(0, max(sample_data$Activated_Percentage)), 
       bty = "l")
  
  # Add the fit line
  abline(fit, col = "black", lwd = 2)
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("R² =", round(fit_summary$r.squared, 3)),
    paste("p-value =", format.pval(fit_summary$coefficients[2,4], digits = 3)),
    paste("Slope =", round(fit_summary$coefficients[2,1], 3)),
    paste("Intercept =", round(fit_summary$coefficients[1,1], 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = "black",
         cex = 1.2,
         bty = "n")
}

# Save the  plot
pdf(file = "../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303/control_all_neurons_plot.pdf",
   width = 6, height = 8)
plot_control(overall_sample_percentages)
dev.off()
```

```{r regression rewards by pct activity in all neurons}
overall_stats <- overall_sample_percentages %>%
  {
    fit <- lm(TotalRewards ~ Activated_Percentage, data = .)
    sum_fit <- summary(fit)
    
    data.frame(
      Analysis = "All_Neurons",
      Slope = coef(fit)[2],
      Intercept = coef(fit)[1],
      p_value = sum_fit$coefficients[2,4],
      r_squared = sum_fit$r.squared
    )
  }

# Write results to CSV
write.csv(overall_stats,
          file = "../results/IEGactivity_WTonly/250303_Rewards_byactivationpercent_allneurons.csv",
          row.names = FALSE)
```

```{r scatter rewards by pct activity in all neurons by sex}
# Calculate overall activation by sample and sex
overall_sample_sex_percentages <- cluster_ieg_sample_df %>%
  group_by(Sample, Sex) %>%
  summarise(
    Total_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    TotalRewards = dplyr::first(TotalRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Total_Cells * 100) %>%
  ungroup()

# Create a plotting function for the control with sex
plot_control_withsex <- function(sample_data) {
  # Fit lines for all data, male, and female
  fit_all <- lm(TotalRewards ~ Activated_Percentage + Sex, data = sample_data)
  fit_male <- lm(TotalRewards ~ Activated_Percentage, data = sample_data[sample_data$Sex == "M", ])
  fit_female <- lm(TotalRewards ~ Activated_Percentage, data = sample_data[sample_data$Sex == "F", ])
  
  # Create the plot
  plot(sample_data$Activated_Percentage, sample_data$TotalRewards, 
       main = "All Neurons",
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(10, max(sample_data$Activated_Percentage)),         
       bty = "l",
       col = ifelse(sample_data$Sex == "M", "#4478AB", "#ED6677"))
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_male, col = "#4478AB", lwd = 2)
  abline(fit_female, col = "#ED6677", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_male <- summary(fit_male)$r.squared
  r2_female <- summary(fit_female)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Male (R² =", round(r2_male, 3), ")"),
    paste("Female (R² =", round(r2_female, 3), ")"),
    paste("Activated % Slope =", round(summary(fit_all)$coefficients[2,1], 3)),
    paste("Sex Slope =", round(summary(fit_all)$coefficients[3,1], 3)),
    paste("Intercept =", round(summary(fit_all)$coefficients[1,1], 3)),
    paste("Activated % p-value =", format.pval(summary(fit_all)$coefficients[2,4], digits = 3)),
    paste("Sex p-value =", format.pval(summary(fit_all)$coefficients[3,4], digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#4478AB", "#ED6677", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")
}

# Save the control plot with sex
pdf(file = "../results/IEGactivity_WTonly/RewardsbyActivatedPercentage_250303/control_all_neurons_plot_bySex.pdf",
    width = 6, height = 8)
plot_control_withsex(overall_sample_sex_percentages)
dev.off()
```

```{r regression rewards by pct activity in all neurons by sex}
overall_sex_stats <- overall_sample_sex_percentages %>%
  {
    # Full model with sex
    fit <- lm(TotalRewards ~ Activated_Percentage + Sex, data = .)
    sum_fit <- summary(fit)
    
    # Sex-specific models
    fit_male <- lm(TotalRewards ~ Activated_Percentage, data = .[.$Sex == "M",])
    fit_female <- lm(TotalRewards ~ Activated_Percentage, data = .[.$Sex == "F",])
    
    data.frame(
      Analysis = "All_Neurons",
      Slope_Activated_Percentage = coef(fit)[2],
      Slope_Sex = coef(fit)[3],
      Intercept = coef(fit)[1],
      p_value_activated_percentage = sum_fit$coefficients[2,4],
      p_value_sex = sum_fit$coefficients[3,4],
      r_squared_full = sum_fit$r.squared,
      r_squared_male = summary(fit_male)$r.squared,
      r_squared_female = summary(fit_female)$r.squared
    )
  }

# Write results to CSV
write.csv(overall_sex_stats, 
          file = "../results/IEGactivity_WTonly/250303_Rewards_byactivationpercent_allneurons_bysex.csv", 
          row.names = FALSE)
```


## Plotting B values
```{r plotting beta coefficients rewards}
# First, go back to the nested data and models to get standard errors
standard_errors <- cluster_ieg_sample_percentages %>%
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(TotalRewards ~ Activated_Percentage, data = .x)),
    std_error = map_dbl(lm_model, ~ summary(.x)$coefficients[2, 2])
  ) %>%
  dplyr::select(Cluster, std_error)

# Join the standard errors with your existing results
cluster_regression_results <- cluster_regression_results %>%
  left_join(standard_errors, by = "Cluster")

# Now create the plot with error bars
p_beta <- ggplot(cluster_regression_results, 
       aes(x = Slope, y = reorder(Cluster, Slope))) +
  geom_bar(stat = "identity", 
           width = 0.7,
           fill = "#888888") +
  geom_errorbar(aes(xmin = Slope - std_error, 
                    xmax = Slope + std_error),
                position = position_dodge(0.7),
                width = 0.2, 
                size = 0.2) +
  labs(x = "Beta Coefficient (Slope)", 
       y = "Cluster") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.y = element_text(color = rep(c("#000000", "#555555"), 
                                 length.out = length(unique(cluster_regression_results$Cluster)))))

p_beta

ggsave('../results/IEGactivity_WTonly/intermed-figs/250225_rewardsbyactivity_betacoeffs.pdf',
       device = pdf,
       width = 6,
       height = 8)
```

```{r selecting sig clusters rewards}
sig_clusters_rewards <- cluster_regression_results[cluster_regression_results$p_value < 0.10, ]

sig_clusters_rewards_names <-sort(as.character(sig_clusters_rewards$Cluster))
print(sig_clusters_rewards)
```

```{r plotting beta coefficients rewards sig only}
sig_clusters_rewards$Cluster <- factor(sig_clusters_rewards$Cluster, 
                                      levels = rev(sort(unique(as.character(sig_clusters_rewards$Cluster)))))

# Now create the plot
p_beta <- ggplot(sig_clusters_rewards, 
       aes(x = Slope, y = Cluster)) +
  geom_bar(stat = "identity", 
           width = 0.7,
           fill = "#888888") +
  geom_errorbar(aes(xmin = Slope - std_error, 
                    xmax = Slope + std_error),
                position = position_dodge(0.7),
                width = 0.2, 
                size = 0.2) +
  labs(x = "Beta Coefficient (Slope)", 
       y = "") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.y = element_text(color = rep(c("#000000", "#555555"), 
                                 length.out = length(unique(sig_clusters_rewards)))))

p_beta

ggsave('../results/IEGactivity_WTonly/250225_rewardsbyactivity_betacoeffs_sigonly.pdf',
  device = pdf,
  width = 5,
  height = 10.5)
```
## Sac Rewards

```{r regression sac rewards by pct activated}
# Fetch expression data for all IEGs
#ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  SacRewards = wt_only$Sac_Total_Rewards,  
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_sample_percentages <- cluster_ieg_sample_df %>%
  group_by(Cluster, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    SacRewards = dplyr::first(SacRewards)  
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

cluster_regression_results <- cluster_ieg_sample_percentages %>%
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(SacRewards ~ Activated_Percentage, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    p_value = map_dbl(summary, ~ .x$coefficients[2, 4]),
    r_squared = map_dbl(summary, ~ .x$r.squared)
  ) %>%
  dplyr::mutate(
    Slope = map_dbl(coefficients, ~ .x[2]),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope, Intercept, p_value, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values <- p.adjust(cluster_regression_results$p_value, method = "BH")

# Add adjusted p-values to the results data frame
cluster_regression_results$AdjustedPValue <- adjusted_p_values
cluster_regression_results <- cluster_regression_results %>%
    arrange(AdjustedPValue)

# Print the updated results
print(cluster_regression_results)

# # # Write the results to a CSV file
write.csv(cluster_regression_results, file = "../results/IEGactivity_WTonly/250303_sac_rewards_byactivationpercent.csv", row.names = FALSE)
```

```{r scatters sac rewards by pct activated}
plot_cluster <- function(cluster, cluster_regression_results, cluster_ieg_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- cluster_regression_results[cluster_regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$SacRewards, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Sac Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l")   
  
  # Fit line
  fit <- lm(SacRewards ~ Activated_Percentage, data = cluster_df)
  
  # Add the fit line
  abline(fit, col = "black", lwd = 2)
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("R² =", round(cluster_results$r_squared, 3)),
    paste("p-value =", format.pval(cluster_results$p_value, digits = 3)),
    paste("Adj. p-value =", format.pval(cluster_results$AdjustedPValue, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = "black",
         cex = 1.2,
         bty = "n")
  
  # Add title with cluster name and coefficients
  title_text <- paste("Cluster", cluster, "\n",
                      "Intercept =", round(cluster_results$Intercept, 3), "\n",
                      "Slope =", round(cluster_results$Slope, 3))
  title(main = title_text, cex.main = 1.5)
}

# Get all unique cluster numbers
clusters <- unique(cluster_regression_results$Cluster)


# Get all unique cluster numbers
clusters <- unique(cluster_regression_results$Cluster)

# Create a directory to save the PDFs if it doesn't exist
dir.create("../results/IEGactivity_WTonly/SacRewardsbyActivatedPercentage_250303", showWarnings = FALSE)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/SacRewardsbyActivatedPercentage_250303/cluster_", cluster, "_plot.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster(cluster, cluster_regression_results, cluster_ieg_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'SacRewardsbyActivatedPercentage' directory.\n")
```

### with Sex
```{r regression sac rewards by pct activated and sex }
# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Sex_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  SacRewards = wt_only$Sac_Total_Rewards, 
  ieg_expression
)
# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_df %>%
  group_by(Cluster, Sex, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    SacRewards = dplyr::first(SacRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Perform linear regression for each cluster
regression_results <- cluster_ieg_Sex_sample_percentages %>%
  # First identify clusters with only one Sex level
  group_by(Cluster) %>%
  mutate(sex_levels_count = n_distinct(Sex)) %>%
  ungroup() %>%
  # Filter to keep only clusters with at least 2 Sex levels
  filter(sex_levels_count >= 2) %>%
  # Now proceed with your analysis on the filtered data
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(SacRewards ~ Activated_Percentage + Sex , data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    r_squared = map_dbl(summary, ~ .x$r.squared),
    p_value_activated_percentage = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 2 && ncol(coef_matrix) >= 4) .x$coefficients[2, 4] else NA_real_
    }),
    p_value_sex = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 3 && ncol(coef_matrix) >= 4) .x$coefficients[3, 4] else NA_real_
    })
  ) %>%
  dplyr::mutate(
    Slope_Activated_Percentage = map_dbl(coefficients, ~ {
      if (length(.x) >= 2) .x[2] else NA_real_
    }),
    Slope_Sex = map_dbl(coefficients, ~ {
      if (length(.x) >= 3) .x[3] else NA_real_
    }),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope_Activated_Percentage, Slope_Sex, Intercept,
         p_value_activated_percentage, p_value_sex, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values_activated <- p.adjust(regression_results$p_value_activated_percentage, method = "BH")
adjusted_p_values_sex <- p.adjust(regression_results$p_value_sex, method = "BH")

# Add adjusted p-values to the results data frame
regression_results$adjusted_p_values_activated <-adjusted_p_values_activated
regression_results$adjusted_p_values_sex <- adjusted_p_values_sex

regression_results <- regression_results %>%
    arrange(adjusted_p_values_activated)

# Print the updated results
print(regression_results)

# # Write the results to a CSV file
write.csv(regression_results, file = "../results/IEGactivity_WTonly/250303_sacrewards_byactivationpercent_bysex.csv", row.names = FALSE)
```

```{r scatters sac rewards by pct activated with sex}
# First, let's create a function for our plotting code
plot_cluster_withsex <- function(cluster, regression_results, cluster_ieg_Sex_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- regression_results[regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_Sex_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$SacRewards, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Sac Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l",
       col = ifelse(cluster_df$Sex == "M", "#4478AB", "#ED6677"))
  
  # Fit lines for all data, male, and female
  fit_all <- lm(SacRewards ~ Activated_Percentage + Sex, data = cluster_df)
  fit_male <- lm(SacRewards ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "M", ])
  fit_female <- lm(SacRewards ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "F", ])
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_male, col = "#4478AB", lwd = 2)
  abline(fit_female, col = "#ED6677", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_male <- summary(fit_male)$r.squared
  r2_female <- summary(fit_female)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Male (R² =", round(r2_male, 3), ")"),
    paste("Female (R² =", round(r2_female, 3), ")"),
    paste("Activated % Slope =", round(cluster_results$Slope_Activated_Percentage, 3)),
    paste("Sex Slope =", round(cluster_results$Slope_Sex, 3)),
    paste("Intercept =", round(cluster_results$Intercept, 3)),
    paste("Activated % p-value =", format.pval(cluster_results$p_value_activated_percentage, digits = 3)),
    paste("Sex p-value =", format.pval(cluster_results$p_value_sex, digits = 3)),
    paste("Activated % Adj. p-value =", format.pval(cluster_results$adjusted_p_values_activated, digits = 3)),
    paste("Sex Adj. p-value =", format.pval(cluster_results$adjusted_p_values_sex, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#4478AB", "#ED6677", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")

} 
# Get all unique cluster numbers
clusters <- unique(regression_results$Cluster)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/SacRewardsbyActivatedPercentage_250303/cluster_", cluster, "_plot_bySex.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster_withsex(cluster, regression_results, cluster_ieg_Sex_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'SacRewardsbyActivatedPercentage' directory.\n")
```

### Control: by Pct Activity in All Clusters

```{r scatter sac rewards by pct activity in all neurons}
# First calculate the activation percentage across all neurons for each sample
overall_sample_percentages <- cluster_ieg_sample_df %>%
  group_by(Sample) %>%
  summarise(
    Total_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    SacRewards = dplyr::first(SacRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Total_Cells * 100) %>%
  ungroup()

# Create a plotting function for the control
plot_control <- function(sample_data) {
  # Fit regression line
  fit <- lm(SacRewards ~ Activated_Percentage, data = sample_data)
  fit_summary <- summary(fit)
  
  # Create the plot
  plot(sample_data$Activated_Percentage, sample_data$SacRewards, 
       main = "All Neurons",
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Sac Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(sample_data$Activated_Percentage)),         
       bty = "l")   
  
  # Add the fit line
  abline(fit, col = "black", lwd = 2)
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("R² =", round(fit_summary$r.squared, 3)),
    paste("p-value =", format.pval(fit_summary$coefficients[2,4], digits = 3)),
    paste("Slope =", round(fit_summary$coefficients[2,1], 3)),
    paste("Intercept =", round(fit_summary$coefficients[1,1], 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = "black",
         cex = 1.2,
         bty = "n")
}

# Save the  plot
pdf(file = "../results/IEGactivity_WTonly/SacRewardsbyActivatedPercentage_250303/control_all_neurons_plot.pdf", 
    width = 6, height = 8)
plot_control(overall_sample_percentages)
dev.off()
```

```{r regression sac rewards by pct activity in all neurons}
overall_stats <- overall_sample_percentages %>%
  {
    fit <- lm(SacRewards ~ Activated_Percentage, data = .)
    sum_fit <- summary(fit)
    
    data.frame(
      Analysis = "All_Neurons",
      Slope = coef(fit)[2],
      Intercept = coef(fit)[1],
      p_value = sum_fit$coefficients[2,4],
      r_squared = sum_fit$r.squared
    )
  }

# Write results to CSV
write.csv(overall_stats,
          file = "../results/IEGactivity_WTonly/250303_sacrewards_byactivationpercent_allneurons.csv",
          row.names = FALSE)
```

```{r scatter sac rewards by pct activity in all neurons by sex}
# Calculate overall activation by sample and sex
overall_sample_sex_percentages <- cluster_ieg_sample_df %>%
  group_by(Sample, Sex) %>%
  summarise(
    Total_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    SacRewards = dplyr::first(SacRewards)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Total_Cells * 100) %>%
  ungroup()

# Create a plotting function for the control with sex
plot_control_withsex <- function(sample_data) {
  # Fit lines for all data, male, and female
  fit_all <- lm(SacRewards ~ Activated_Percentage + Sex, data = sample_data)
  fit_male <- lm(SacRewards ~ Activated_Percentage, data = sample_data[sample_data$Sex == "M", ])
  fit_female <- lm(SacRewards ~ Activated_Percentage, data = sample_data[sample_data$Sex == "F", ])
  
  # Create the plot
  plot(sample_data$Activated_Percentage, sample_data$SacRewards, 
       main = "All Neurons",
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Sac Rewards", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(sample_data$Activated_Percentage)),         
       bty = "l",
       col = ifelse(sample_data$Sex == "M", "#4478AB", "#ED6677"))
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_male, col = "#4478AB", lwd = 2)
  abline(fit_female, col = "#ED6677", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_male <- summary(fit_male)$r.squared
  r2_female <- summary(fit_female)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Male (R² =", round(r2_male, 3), ")"),
    paste("Female (R² =", round(r2_female, 3), ")"),
    paste("Activated % Slope =", round(summary(fit_all)$coefficients[2,1], 3)),
    paste("Sex Slope =", round(summary(fit_all)$coefficients[3,1], 3)),
    paste("Intercept =", round(summary(fit_all)$coefficients[1,1], 3)),
    paste("Activated % p-value =", format.pval(summary(fit_all)$coefficients[2,4], digits = 3)),
    paste("Sex p-value =", format.pval(summary(fit_all)$coefficients[3,4], digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#4478AB", "#ED6677", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")
}

# Save the control plot with sex
pdf(file = "../results/IEGactivity_WTonly/SacRewardsbyActivatedPercentage_250303/control_all_neurons_plot_bySex.pdf",
    width = 6, height = 8)
plot_control_withsex(overall_sample_sex_percentages)
dev.off()
```

```{r get overall sex stats}
overall_sex_stats <- overall_sample_sex_percentages %>%
  {
    # Full model with sex
    fit <- lm(SacRewards ~ Activated_Percentage + Sex, data = .)
    sum_fit <- summary(fit)
    
    # Sex-specific models
    fit_male <- lm(SacRewards ~ Activated_Percentage, data = .[.$Sex == "M",])
    fit_female <- lm(SacRewards ~ Activated_Percentage, data = .[.$Sex == "F",])
    
    data.frame(
      Analysis = "All_Neurons",
      Slope_Activated_Percentage = coef(fit)[2],
      Slope_Sex = coef(fit)[3],
      Intercept = coef(fit)[1],
      p_value_activated_percentage = sum_fit$coefficients[2,4],
      p_value_sex = sum_fit$coefficients[3,4],
      r_squared_full = sum_fit$r.squared,
      r_squared_male = summary(fit_male)$r.squared,
      r_squared_female = summary(fit_female)$r.squared
    )
  }

# Write results to CSV
write.csv(overall_sex_stats, 
          file = "../results/IEGactivity_WTonly/250303_sacrewards_byactivationpercent_allneurons_bysex.csv", 
          row.names = FALSE)
```

## Distance
```{r regressions distance by pct activated}
# Fetch expression data for all IEGs
#ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  Distance = wt_only$Avg_Test_Distance,  
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_sample_percentages <- cluster_ieg_sample_df %>%
  group_by(Cluster, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    Distance = dplyr::first(Distance)  
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Removing outlier for distance
cluster_ieg_sample_percentages <- cluster_ieg_sample_percentages %>%
  filter(Sample != "Het_F_6")

# Perform linear regression for each cluster
cluster_regression_results <- cluster_ieg_sample_percentages %>%
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(Distance ~ Activated_Percentage, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    p_value = map_dbl(summary, ~ .x$coefficients[2, 4]),
    r_squared = map_dbl(summary, ~ .x$r.squared)
  ) %>%
  dplyr::mutate(
    Slope = map_dbl(coefficients, ~ .x[2]),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope, Intercept, p_value, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values <- p.adjust(cluster_regression_results$p_value, method = "BH")

# Add adjusted p-values to the results data frame
cluster_regression_results$AdjustedPValue <- adjusted_p_values
cluster_regression_results <- cluster_regression_results %>%
    arrange(AdjustedPValue)

# Print the updated results
print(cluster_regression_results)

# # Write the results to a CSV file
write.csv(cluster_regression_results, file = "../results/IEGactivity_WTonly/250303_Distance_byactivationpercent.csv", row.names = FALSE)
```

```{r scatters distance by pct activated}
# First, let's create a function for our plotting code
plot_cluster <- function(cluster, cluster_regression_results, cluster_ieg_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- cluster_regression_results[cluster_regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$Distance, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Total Distance", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l")   
  
  # Fit line
  fit <- lm(Distance ~ Activated_Percentage, data = cluster_df)
  
  # Add the fit line
  abline(fit, col = "black", lwd = 2)
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("R² =", round(cluster_results$r_squared, 3)),
    paste("p-value =", format.pval(cluster_results$p_value, digits = 3)),
    paste("Adj. p-value =", format.pval(cluster_results$AdjustedPValue, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = "black",
         cex = 1.2,
         bty = "n")
  
  # Add title with cluster name and coefficients
  title_text <- paste("Cluster", cluster, "\n",
                      "Intercept =", round(cluster_results$Intercept, 3), "\n",
                      "Slope =", round(cluster_results$Slope, 3))
  title(main = title_text, cex.main = 1.5)
}

# Get all unique cluster numbers
clusters <- unique(cluster_regression_results$Cluster)

# Create a directory to save the PDFs if it doesn't exist
dir.create("../results/IEGactivity_WTonly/DistancebyActivatedPercentage_250303", showWarnings = FALSE)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/DistancebyActivatedPercentage_250303/cluster_", cluster, "_plot.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster(cluster, cluster_regression_results, cluster_ieg_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'DistancebyActivatedPercentage' directory.\n")
```

### with Sex
```{r regressions distance by pct activated and sex}
# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Sex_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  Distance = wt_only$Avg_Test_Distance, 
  ieg_expression
)
# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_df %>%
  group_by(Cluster, Sex, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    Distance = dplyr::first(Distance)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Removing outlier for distance
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_percentages %>%
  filter(Sample != "Het_F_6")

# Perform linear regression for each cluster, skipping those with only one Sex level
regression_results <- cluster_ieg_Sex_sample_percentages %>%
  # First identify clusters with only one Sex level
  group_by(Cluster) %>%
  mutate(sex_levels_count = n_distinct(Sex)) %>%
  ungroup() %>%
  # Filter to keep only clusters with at least 2 Sex levels
  filter(sex_levels_count >= 2) %>%
  # Now proceed with your analysis on the filtered data
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(Distance ~ Activated_Percentage + Sex, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    p_value_activated_percentage = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 2 && ncol(coef_matrix) >= 4) .x$coefficients[2, 4] else NA_real_
    }),
    p_value_sex = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 3 && ncol(coef_matrix) >= 4) .x$coefficients[3, 4] else NA_real_
    })
  ) %>%
  dplyr::mutate(
    Slope_Activated_Percentage = map_dbl(coefficients, ~ {
      if (length(.x) >= 2) .x[2] else NA_real_
    }),
    Slope_Sex = map_dbl(coefficients, ~ {
      if (length(.x) >= 3) .x[3] else NA_real_
    }),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope_Activated_Percentage, Slope_Sex, Intercept,
         p_value_activated_percentage, p_value_sex)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values_activated <- p.adjust(regression_results$p_value_activated_percentage, method = "BH")
adjusted_p_values_sex <- p.adjust(regression_results$p_value_sex, method = "BH")

# Add adjusted p-values to the results data frame
regression_results$adjusted_p_values_activated <-adjusted_p_values_activated
regression_results$adjusted_p_values_sex <- adjusted_p_values_sex

regression_results <- regression_results %>%
    arrange(adjusted_p_values_activated)

# Print the updated results
print(regression_results)

# Write the results to a CSV file
write.csv(regression_results, file = "../results/IEGactivity_WTonly/250303_distance_byactivationpercent_bysex.csv", row.names = FALSE)
```

```{r scatters distance by pct activated with sex}
# First, let's create a function for our plotting code
plot_cluster_withsex <- function(cluster, regression_results, cluster_ieg_Sex_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- regression_results[regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_Sex_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$Distance, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Mean Distance", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l",
       col = ifelse(cluster_df$Sex == "M", "#4478AB", "#ED6677"))
  
  # Fit lines for all data, male, and female
  fit_all <- lm(Distance ~ Activated_Percentage + Sex, data = cluster_df)
  fit_male <- lm(Distance ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "M", ])
  fit_female <- lm(Distance ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "F", ])
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_male, col = "#4478AB", lwd = 2)
  abline(fit_female, col = "#ED6677", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_male <- summary(fit_male)$r.squared
  r2_female <- summary(fit_female)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Male (R² =", round(r2_male, 3), ")"),
    paste("Female (R² =", round(r2_female, 3), ")"),
    paste("Activated % Slope =", round(cluster_results$Slope_Activated_Percentage, 3)),
    paste("Sex Slope =", round(cluster_results$Slope_Sex, 3)),
    paste("Intercept =", round(cluster_results$Intercept, 3)),
    paste("Activated % p-value =", format.pval(cluster_results$p_value_activated_percentage, digits = 3)),
    paste("Sex p-value =", format.pval(cluster_results$p_value_sex, digits = 3)),
    paste("Activated % Adj. p-value =", format.pval(cluster_results$adjusted_p_values_activated, digits = 3)),
    paste("Sex Adj. p-value =", format.pval(cluster_results$adjusted_p_values_sex, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#4478AB", "#ED6677", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")

} 
# Get all unique cluster numbers
clusters <- unique(regression_results$Cluster)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/DistancebyActivatedPercentage_250303/cluster_", cluster, "_plot_bySex.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster_withsex(cluster, regression_results, cluster_ieg_Sex_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'DistancebyActivatedPercentage' directory.\n")
```

## Sac Distance

```{r regressions sac distance by pct activated}
# Fetch expression data for all IEGs
#ieg_expression <- FetchData(wt_only, vars = filtered_iegs)

# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  SacDistance = wt_only$Sac_Test_Distance,
  ieg_expression
)

# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_sample_percentages <- cluster_ieg_sample_df %>%
  group_by(Cluster, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    SacDistance = dplyr::first(SacDistance) 
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Removing outlier for distance
cluster_ieg_sample_percentages <- cluster_ieg_sample_percentages %>%
  filter(Sample != "Het_F_6")

cluster_regression_results <- cluster_ieg_sample_percentages %>%
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(SacDistance ~ Activated_Percentage, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    p_value = map_dbl(summary, ~ .x$coefficients[2, 4]),
    r_squared = map_dbl(summary, ~ .x$r.squared)
  ) %>%
  dplyr::mutate(
    Slope = map_dbl(coefficients, ~ .x[2]),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope, Intercept, p_value, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values <- p.adjust(cluster_regression_results$p_value, method = "BH")

# Add adjusted p-values to the results data frame
cluster_regression_results$AdjustedPValue <- adjusted_p_values
cluster_regression_results <- cluster_regression_results %>%
    arrange(AdjustedPValue)

# Print the updated results
print(cluster_regression_results)

# # # Write the results to a CSV file
write.csv(cluster_regression_results, file = "../results/IEGactivity_WTonly/250303_sac_distance_byactivationpercent.csv", row.names = FALSE)
```

```{r scatters sac distance by pct activated}
# First, let's create a function for our plotting code
plot_cluster <- function(cluster, cluster_regression_results, cluster_ieg_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- cluster_regression_results[cluster_regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$SacDistance, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Sac Distance", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l")   
  
  # Fit line
  fit <- lm(SacDistance ~ Activated_Percentage, data = cluster_df)
  
  # Add the fit line
  abline(fit, col = "black", lwd = 2)
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("R² =", round(cluster_results$r_squared, 3)),
    paste("p-value =", format.pval(cluster_results$p_value, digits = 3)),
    paste("Adj. p-value =", format.pval(cluster_results$AdjustedPValue, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = "black",
         cex = 1.2,
         bty = "n")
  
  # Add title with cluster name and coefficients
  title_text <- paste("Cluster", cluster, "\n",
                      "Intercept =", round(cluster_results$Intercept, 3), "\n",
                      "Slope =", round(cluster_results$Slope, 3))
  title(main = title_text, cex.main = 1.5)
}

# Get all unique cluster numbers
clusters <- unique(cluster_regression_results$Cluster)

# Create a directory to save the PDFs if it doesn't exist
dir.create("../results/IEGactivity_WTonly/Sac_DistancebyActivatedPercentage_250303", showWarnings = FALSE)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/Sac_DistancebyActivatedPercentage_250303/cluster_", cluster, "_plot.pdf"), width = 6, height = 8)
  
  # Create the plot
  plot_cluster(cluster, cluster_regression_results, cluster_ieg_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'Sac_DistancebyActivatedPercentage' directory.\n")
```

### with Sex
```{r regression sac distance by pct activated and sex}
# Create a data frame with cluster assignments, Sex, sample, and IEG expression
cluster_ieg_Sex_sample_df <- data.frame(
  Cluster = Idents(wt_only),
  Sex = wt_only$sex,
  Sample = wt_only$sample,
  SacDistance = wt_only$Sac_Test_Distance, 
  ieg_expression
)
# Calculate the percentage of cells expressing 2 or more IEGs within each cluster, Sex, and sample
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_df %>%
  group_by(Cluster, Sex, Sample) %>%
  summarise(
    Cluster_Cells = n(),
    Activated_Cells = sum(rowSums(across(all_of(filtered_iegs)) > 0) >= 2),
    SacDistance = dplyr::first(SacDistance)
  ) %>%
  dplyr::mutate(Activated_Percentage = Activated_Cells / Cluster_Cells * 100) %>%
  ungroup()

# Removing outlier for distance
cluster_ieg_Sex_sample_percentages <- cluster_ieg_Sex_sample_percentages %>%
  filter(Sample != "Het_F_6")

# Perform linear regression for each cluster, skipping those with only one Sex level
regression_results <- cluster_ieg_Sex_sample_percentages %>%
  # First identify clusters with only one Sex level
  group_by(Cluster) %>%
  mutate(sex_levels_count = n_distinct(Sex)) %>%
  ungroup() %>%
  # Filter to keep only clusters with at least 2 Sex levels
  filter(sex_levels_count >= 2) %>%
  # Now proceed with your analysis on the filtered data
  group_by(Cluster) %>%
  nest() %>%
  dplyr::mutate(
    lm_model = map(data, ~ lm(SacDistance ~ Activated_Percentage + Sex, data = .x)),
    summary = map(lm_model, summary),
    coefficients = map(lm_model, coef),
    r_squared = map_dbl(summary, ~ .x$r.squared),
    p_value_activated_percentage = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 2 && ncol(coef_matrix) >= 4) .x$coefficients[2, 4] else NA_real_
    }),
    p_value_sex = map_dbl(summary, ~ {
      coef_matrix <- .x$coefficients
      if (nrow(coef_matrix) >= 3 && ncol(coef_matrix) >= 4) .x$coefficients[3, 4] else NA_real_
    })
  ) %>%
  dplyr::mutate(
    Slope_Activated_Percentage = map_dbl(coefficients, ~ {
      if (length(.x) >= 2) .x[2] else NA_real_
    }),
    Slope_Sex = map_dbl(coefficients, ~ {
      if (length(.x) >= 3) .x[3] else NA_real_
    }),
    Intercept = map_dbl(coefficients, ~ .x[1])
  ) %>%
  dplyr::select(Cluster, Slope_Activated_Percentage, Slope_Sex, Intercept,
         p_value_activated_percentage, p_value_sex, r_squared)

# Adjust p-values for multiple comparisons (optional)
adjusted_p_values_activated <- p.adjust(regression_results$p_value_activated_percentage, method = "BH")
adjusted_p_values_sex <- p.adjust(regression_results$p_value_sex, method = "BH")

# Add adjusted p-values to the results data frame
regression_results$adjusted_p_values_activated <-adjusted_p_values_activated
regression_results$adjusted_p_values_sex <- adjusted_p_values_sex

regression_results <- regression_results %>%
    arrange(adjusted_p_values_activated)

# Print the updated results
print(regression_results)

# # Write the results to a CSV file
write.csv(regression_results, file = "../results/IEGactivity_WTonly/250303_SacDistance_byactivationpercent_bysex.csv", row.names = FALSE)
```

```{r scatters sac distance by pct activated with sex}
# First, let's create a function for our plotting code
plot_cluster_withsex <- function(cluster, regression_results, cluster_ieg_Sex_sample_percentages) {
  # Extract the results for the specific cluster we're plotting
  cluster_results <- regression_results[regression_results$Cluster == cluster, ]
  
  # Filter the data for the specific cluster
  cluster_df <- cluster_ieg_Sex_sample_percentages %>% 
    filter(Cluster == cluster)
  
  # Create the plot
  plot(cluster_df$Activated_Percentage, cluster_df$SacDistance, 
       main = paste("Cluster", cluster),
       pch = 16, cex = 2, 
       xlab = "Activated Percentage", ylab = "Sac Distance", 
       cex.lab = 1.5, cex.axis = 1.5,
       cex.main = 2,        
       xlim = c(0, max(cluster_df$Activated_Percentage)),         
       bty = "l",
       col = ifelse(cluster_df$Sex == "M", "#4478AB", "#ED6677"))
  
  # Fit lines for all data, male, and female
  fit_all <- lm(SacDistance ~ Activated_Percentage + Sex, data = cluster_df)
  fit_male <- lm(SacDistance ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "M", ])
  fit_female <- lm(SacDistance ~ Activated_Percentage, data = cluster_df[cluster_df$Sex == "F", ])
  
  # Add the fit lines
  abline(fit_all, col = "black", lwd = 2)
  abline(fit_male, col = "#4478AB", lwd = 2)
  abline(fit_female, col = "#ED6677", lwd = 2)
  
  # Calculate R-squared values for each fit
  r2_all <- summary(fit_all)$r.squared
  r2_male <- summary(fit_male)$r.squared
  r2_female <- summary(fit_female)$r.squared
  
  # Create legend labels with statistics
  legend_labels <- c(
    paste("All (R² =", round(r2_all, 3), ")"),
    paste("Male (R² =", round(r2_male, 3), ")"),
    paste("Female (R² =", round(r2_female, 3), ")"),
    paste("Activated % Slope =", round(cluster_results$Slope_Activated_Percentage, 3)),
    paste("Sex Slope =", round(cluster_results$Slope_Sex, 3)),
    paste("Intercept =", round(cluster_results$Intercept, 3)),
    paste("Activated % p-value =", format.pval(cluster_results$p_value_activated_percentage, digits = 3)),
    paste("Sex p-value =", format.pval(cluster_results$p_value_sex, digits = 3)),
    paste("Activated % Adj. p-value =", format.pval(cluster_results$adjusted_p_values_activated, digits = 3)),
    paste("Sex Adj. p-value =", format.pval(cluster_results$adjusted_p_values_sex, digits = 3))
  )
  
  # Add legend
  legend("topleft", 
         legend = legend_labels,
         col = c("black", "#4478AB", "#ED6677", rep("black", 5)),
         lty = c(1, 1, 1, rep(NA, 5)),
         cex = 1.2,
         bty = "n")

} 


# Get all unique cluster numbers
clusters <- unique(regression_results$Cluster)

# Iterate through clusters
for (cluster in clusters) {
  # Open a PDF device
  pdf(file = paste0("../results/IEGactivity_WTonly/Sac_DistancebyActivatedPercentage_250303/cluster_", cluster, "_plot_bySex.pdf"), width = 6, height = 8)
  # Create the plot
  plot_cluster_withsex(cluster, regression_results, cluster_ieg_Sex_sample_percentages)
  
  # Close the PDF device
  dev.off()
  
  # Print progress
  cat("Saved plot for cluster", cluster, "\n")
}

cat("All plots have been saved in the 'SacDistancebyActivatedPercentage' directory.\n")
```

# Gene Module Analysis

```{r add module scores}
ieg_lst <- list(ieg_62 = filtered_iegs)

wt_only <- AddModuleScore(
  object = wt_only,
  features = ieg_lst,
  pool = NULL,
  nbin = 24,
  ctrl = 100,
  k = FALSE,
  assay = NULL,
  name = "IEG_score",
  seed = 1,
  search = FALSE,
  slot = "data")
```

```{r set up dataframe}
sample_info = wt_only@meta.data %>% 
  dplyr::select(
  sample, 
  rewards = Avg_Total_Rewards, 
  sac_rewards = Sac_Total_Rewards,
  distance = Avg_Test_Distance,
  sac_distance = Sac_Test_Distance,
  cell_type, 
  genotype, 
  sex, 
  score_ieg62 = IEG_score1)
```

## Rewards

```{r linear regression for IEG module rewards}
model_summary_df <- data.frame()

for (cluster in unique(sample_info$cell_type)) {
  # Filter the data frame for the current cluster
  cluster_df <- sample_info %>% 
    filter(cell_type == cluster) %>% 
    distinct
  
  # Fit the linear regression model
  lm_model <- lm(rewards ~ score_ieg62, data = cluster_df)
  
  # Extract the model summary
  model_summary <- summary(lm_model)
  
  # Create a named vector with model summary statistics
  model_summary_vec <- c(
    `(Intercept)` = coef(lm_model)[1],
    `(Intercept) p-value` = coef(model_summary)[1, "Pr(>|t|)"],
    coef(lm_model)[2],
    `Std. Error` = coef(model_summary)[2, "Std. Error"],
    `t_value` = coef(model_summary)[2, "t value"],
    `p_value` = coef(model_summary)[2, "Pr(>|t|)"],
    r_squared = model_summary$r.squared,
    adj_r_squared = model_summary$adj.r.squared,
    f_statistic = model_summary$fstatistic[1],
    f_pvalue = pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)
  )
  
  # Convert the named vector to a data frame
  model_summary_vec_df <- data.frame(t(model_summary_vec))
  
  # Add a column for the cluster
  model_summary_vec_df$Cluster <- cluster
  
  # Append the data frame to the overall result
  model_summary_df <- rbind(model_summary_df, model_summary_vec_df)
}

model_summary_df$adj_p_value <- p.adjust(model_summary_df$p_value, method = 'BH')

model_summary_df %>% 
  write_csv('../results/IEGactivity_WTonly/250303_genemodulescore_results_rewards.csv')
```

```{r plots for rewards}
# filter only significant results
sig_df <- model_summary_df %>% filter(p_value < 0.05)

cts <- sig_df$Cluster %>% unique()

dir.create("../results/IEGactivity_WTonly/RewardsbyModuleScore_Figs", showWarnings = FALSE)

for (ct in cts){
  plot_df <- sample_info %>% 
    filter(cell_type == ct)
  
    pp <- ggplot(plot_df, aes(score_ieg62, rewards)) +
    geom_smooth(method = 'lm', se = FALSE, fullrange = TRUE, color = 'black', linewidth = 0.75) +
    geom_smooth(method = 'lm', aes(color = sex), se = FALSE, fullrange = TRUE, linewidth = 0.75) +
    scale_color_manual(values = palette_sex) +
    scale_x_continuous(expand = expansion(c(0, 0))) +
    labs(title = ct, color = 'Sex', x = 'Score_IEG62', y = 'Rewards') +
    theme_bw() +
    theme(panel.grid = element_blank(),
        plot.title = element_text(size = 16),
        axis.text = element_text(size = 12, color = 'black'),
        axis.title = element_text(size = 14))
    
    print(pp)
    ct_name <- gsub('/', '-', ct)
    ggsave(paste0('../results/IEGactivity_WTonly/RewardsbyModuleScore_Figs/250303_linearplot_', ct_name, '_genemodulescore.pdf'), pp, width = 5, height = 3.5)
  }
```

## Sac rewards

```{r linear regression for IEG module sac_rewards}
model_summary_df <- data.frame()

for (cluster in unique(sample_info$cell_type)) {
  # Filter the data frame for the current cluster
  cluster_df <- sample_info %>% 
    filter(cell_type == cluster) %>% 
    distinct
  
  # Fit the linear regression model
  lm_model <- lm(sac_rewards ~ score_ieg62, data = cluster_df)
  
  # Extract the model summary
  model_summary <- summary(lm_model)
  
  # Create a named vector with model summary statistics
  model_summary_vec <- c(
    `(Intercept)` = coef(lm_model)[1],
    `(Intercept) p-value` = coef(model_summary)[1, "Pr(>|t|)"],
    coef(lm_model)[2],
    `Std. Error` = coef(model_summary)[2, "Std. Error"],
    `t_value` = coef(model_summary)[2, "t value"],
    `p_value` = coef(model_summary)[2, "Pr(>|t|)"],
    r_squared = model_summary$r.squared,
    adj_r_squared = model_summary$adj.r.squared,
    f_statistic = model_summary$fstatistic[1],
    f_pvalue = pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)
  )
  
  # Convert the named vector to a data frame
  model_summary_vec_df <- data.frame(t(model_summary_vec))
  
  # Add a column for the cluster
  model_summary_vec_df$Cluster <- cluster
  
  # Append the data frame to the overall result
  model_summary_df <- rbind(model_summary_df, model_summary_vec_df)
}

model_summary_df$adj_p_value <- p.adjust(model_summary_df$p_value, method = 'BH')

model_summary_df %>% 
  write_csv('../results/IEGactivity_WTonly/250303_genemodulescore_results_sac_rewards.csv')
```

```{r plots for sac_rewards}
# filter only significant results
sig_df <- model_summary_df %>% filter(p_value < 0.05)

cts <- sig_df$Cluster %>% unique()

dir.create("../results/IEGactivity_WTonly/sac_rewardsbyModuleScore_Figs", showWarnings = FALSE)

for (ct in cts){
  plot_df <- sample_info %>% 
    filter(cell_type == ct)
  
    pp <- ggplot(plot_df, aes(score_ieg62, sac_rewards)) +
    geom_smooth(method = 'lm', se = FALSE, fullrange = TRUE, color = 'black', linewidth = 0.75) +
    geom_smooth(method = 'lm', aes(color = sex), se = FALSE, fullrange = TRUE, linewidth = 0.75) +
    scale_color_manual(values = palette_sex) +
    scale_x_continuous(expand = expansion(c(0, 0))) +
    labs(title = ct, color = 'Sex', x = 'Score_IEG62', y = 'sac_rewards') +
    theme_bw() +
    theme(panel.grid = element_blank(),
        plot.title = element_text(size = 16),
        axis.text = element_text(size = 12, color = 'black'),
        axis.title = element_text(size = 14))
    
    print(pp)
    ct_name <- gsub('/', '-', ct)
    ggsave(paste0('../results/IEGactivity_WTonly/sac_rewardsbyModuleScore_Figs/250303_linearplot_', ct_name, '_genemodulescore.pdf'), pp, width = 5, height = 3.5)
  }
```

## Distance

```{r linear regression for IEG module distance}
model_summary_df <- data.frame()

for (cluster in unique(sample_info$cell_type)) {
  # Filter the data frame for the current cluster
  cluster_df <- sample_info %>% 
    filter(cell_type == cluster) %>% 
    distinct
  
  # Fit the linear regression model
  lm_model <- lm(distance ~ score_ieg62, data = cluster_df)
  
  # Extract the model summary
  model_summary <- summary(lm_model)
  
  # Create a named vector with model summary statistics
  model_summary_vec <- c(
    `(Intercept)` = coef(lm_model)[1],
    `(Intercept) p-value` = coef(model_summary)[1, "Pr(>|t|)"],
    coef(lm_model)[2],
    `Std. Error` = coef(model_summary)[2, "Std. Error"],
    `t_value` = coef(model_summary)[2, "t value"],
    `p_value` = coef(model_summary)[2, "Pr(>|t|)"],
    r_squared = model_summary$r.squared,
    adj_r_squared = model_summary$adj.r.squared,
    f_statistic = model_summary$fstatistic[1],
    f_pvalue = pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)
  )
  
  # Convert the named vector to a data frame
  model_summary_vec_df <- data.frame(t(model_summary_vec))
  
  # Add a column for the cluster
  model_summary_vec_df$Cluster <- cluster
  
  # Append the data frame to the overall result
  model_summary_df <- rbind(model_summary_df, model_summary_vec_df)
}

model_summary_df$adj_p_value <- p.adjust(model_summary_df$p_value, method = 'BH')

model_summary_df %>% 
  write_csv('../results/IEGactivity_WTonly/250303_genemodulescore_results_distance.csv')
```

```{r plots for distance}
# filter only significant results
sig_df <- model_summary_df %>% filter(p_value < 0.05)

cts <- sig_df$Cluster %>% unique()

dir.create("../results/IEGactivity_WTonly/distancebyModuleScore_Figs", showWarnings = FALSE)

for (ct in cts){
  plot_df <- sample_info %>% 
    filter(cell_type == ct)
  
    pp <- ggplot(plot_df, aes(score_ieg62, distance)) +
    geom_smooth(method = 'lm', se = FALSE, fullrange = TRUE, color = 'black', linewidth = 0.75) +
    geom_smooth(method = 'lm', aes(color = sex), se = FALSE, fullrange = TRUE, linewidth = 0.75) +
    scale_color_manual(values = palette_sex) +
    scale_x_continuous(expand = expansion(c(0, 0))) +
    labs(title = ct, color = 'Sex', x = 'Score_IEG62', y = 'distance') +
    theme_bw() +
    theme(panel.grid = element_blank(),
        plot.title = element_text(size = 16),
        axis.text = element_text(size = 12, color = 'black'),
        axis.title = element_text(size = 14))
    
    print(pp)
    ct_name <- gsub('/', '-', ct)
    ggsave(paste0('../results/IEGactivity_WTonly/distancebyModuleScore_Figs/250303_linearplot_', ct_name, '_genemodulescore.pdf'), pp, width = 5, height = 3.5)
  }
```

## Sac distance

```{r linear regression for IEG module sac_distance}
model_summary_df <- data.frame()

for (cluster in unique(sample_info$cell_type)) {
  # Filter the data frame for the current cluster
  cluster_df <- sample_info %>% 
    filter(cell_type == cluster) %>% 
    distinct
  
  # Fit the linear regression model
  lm_model <- lm(sac_distance ~ score_ieg62, data = cluster_df)
  
  # Extract the model summary
  model_summary <- summary(lm_model)
  
  # Create a named vector with model summary statistics
  model_summary_vec <- c(
    `(Intercept)` = coef(lm_model)[1],
    `(Intercept) p-value` = coef(model_summary)[1, "Pr(>|t|)"],
    coef(lm_model)[2],
    `Std. Error` = coef(model_summary)[2, "Std. Error"],
    `t_value` = coef(model_summary)[2, "t value"],
    `p_value` = coef(model_summary)[2, "Pr(>|t|)"],
    r_squared = model_summary$r.squared,
    adj_r_squared = model_summary$adj.r.squared,
    f_statistic = model_summary$fstatistic[1],
    f_pvalue = pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)
  )
  
  # Convert the named vector to a data frame
  model_summary_vec_df <- data.frame(t(model_summary_vec))
  
  # Add a column for the cluster
  model_summary_vec_df$Cluster <- cluster
  
  # Append the data frame to the overall result
  model_summary_df <- rbind(model_summary_df, model_summary_vec_df)
}

model_summary_df$adj_p_value <- p.adjust(model_summary_df$p_value, method = 'BH')

model_summary_df %>% 
  write_csv('../results/IEGactivity_WTonly/250303_genemodulescore_results_sac_distance.csv')
```

```{r plots for sac_distance}
# filter only significant results
sig_df = model_summary_df %>% filter(p_value < 0.05)

cts = sig_df$Cluster %>% unique()

dir.create("../results/IEGactivity_WTonly/sac_distancebyModuleScore_Figs", showWarnings = FALSE)

for (ct in cts){
  plot_df = sample_info %>% 
    filter(cell_type == ct)
  
    pp <- ggplot(plot_df, aes(score_ieg62, sac_distance)) +
    geom_smooth(method = 'lm', se = FALSE, fullrange = TRUE, color = 'black', linewidth = 0.75) +
    geom_smooth(method = 'lm', aes(color = sex), se = FALSE, fullrange = TRUE, linewidth = 0.75) +
    scale_color_manual(values = palette_sex) +
    scale_x_continuous(expand = expansion(c(0, 0))) +
    labs(title = ct, color = 'Sex', x = 'Score_IEG62', y = 'sac_distance') +
    theme_bw() +
    theme(panel.grid = element_blank(),
        plot.title = element_text(size = 16),
        axis.text = element_text(size = 12, color = 'black'),
        axis.title = element_text(size = 14))
    
    print(pp)
    ct_name = gsub('/', '-', ct)
    ggsave(paste0('../results/IEGactivity_WTonly/sac_distancebyModuleScore_Figs/250303_linearplot_', ct_name, '_genemodulescore.pdf'), pp, width = 5, height = 3.5)
  }
```

# End
